---
title: "measure-bias"
format: html
editor: visual
---

# Import libraries

```{r}
library(dplyr)
library(sf)
library(ggplot2)
library(tmap)
library(viridis)
library(spdep)
library(sysfonts)
library(showtextdb)
library(classInt)
library(scales)
library(forcats)
library(spgwr)
library(energy)
```

# Set themes

```{r}
#| include = FALSE
source("./style/data-visualisation_theme.R")
```

# Set parameters

```{r}
# Set digital trace data source: twitter, fb_tts_census_week, fb_tts_census_month, fb_stt_census_week, fb_stt_census_month, mapp1, mapp2
dfd <- "mapp2"

# Set weights scheme: queen, fbw, knn, db
w_scheme <- 'queen'
```

# Data import

## Census and boundaries

```{r}
wd <- "/Volumes/rdm04/DEBIAS"
wd_local <- "/Users/carmen/Documents/github/de-bias"
```

```{r}
if (dfd == "twitter") {
  output_data_dir <- paste0(wd, "/data/outputs/twitter/monthly/populations/active-population-bias")
  output_fig_dir <- paste0(wd_local, "/bias-detection/outputs/preliminary/twitter/size-bias")
} else if (dfd == "fb_tts_census_week") {
  output_data_dir <- paste0(wd, "/data/outputs/fb/census-week/populations/tts/active-population-bias")
  output_fig_dir <- paste0(wd_local, "/bias-detection/outputs/preliminary/fb_tts/size-bias")
} else if (dfd == "fb_tts_census_month") {
  output_data_dir <- paste0(wd, "/data/outputs/fb/census-month/populations/tts/active-population-bias")
  output_fig_dir <- paste0(wd_local, "/bias-detection/outputs/preliminary/fb_tts/size-bias")
} else if (dfd == "fb_stt_census_week") {
  output_data_dir <- paste0(wd, "/data/outputs/fb/census-week/populations/stt/active-population-bias")
  output_fig_dir <- paste0(wd_local, "/bias-detection/outputs/preliminary/fb_stt/size-bias")
} else if (dfd == "fb_stt_census_month") {
  output_data_dir <- paste0(wd, "/data/outputs/fb/census-month/populations/stt/active-population-bias")
  output_fig_dir <- paste0(wd_local, "/bias-detection/outputs/preliminary/fb_stt/size-bias")
} else if (dfd == "mapp1") {
  output_data_dir <- paste0(wd, "/data/outputs/mapp1/census-month/populations/active-population-bias")
  output_fig_dir <- paste0(wd_local, "/bias-detection/outputs/preliminary/mapp1/size-bias")
} else if (dfd == "mapp2") {
  output_data_dir <- paste0(wd, "/data/outputs/mapp2/census-month/populations/active-population-bias")
  output_fig_dir <- paste0(wd_local, "/bias-detection/outputs/preliminary/mapp2/size-bias")
}

dir.create(file.path(output_data_dir, "base-case"), recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(output_fig_dir, "base-case"), recursive = TRUE, showWarnings = FALSE)
```

```{r}
df_pop_census <- read.csv(paste0(wd, "/data/inputs/census/census2021-ts/census2021-ts001/census2021-ts001-ltla.csv")) %>% select(date, geography, geography.code, Residence.type..Total..measures..Value) %>% rename("code" = "geography.code", "name" = "geography", "pop" = "Residence.type..Total..measures..Value")

df_boundaries <- st_read(paste0(wd, "/data/inputs/geographies/boundaries/LAD_Dec_2021_GB_BFC_2022.gpkg")) %>% select(LAD21CD, SHAPE) %>% rename("code" = "LAD21CD")

df_boundaries_up <- st_read(paste0(wd, "/data/inputs/geographies/boundaries/RGN_Dec_2021_EN_BFC_2022.gpkg")) %>% st_simplify(preserveTopology = FALSE, dTolerance = 1000)

df_boundaries_hex <- st_read(paste0(wd, "/data/inputs/geographies/hexboundaries/uk-local-authority-districts-2021.geojson")) %>% st_transform(st_crs(df_boundaries_up)) %>% rename("code" = "id")
```

## Digital trace data

This first line defines the data source to be analysed. We have five options:

-   `twitter`

-   `fb_tts_census_week`

-   `fb_tts_census_month`

-   `fb_stt_census_week`

-   `fb_stt_census_month`

-   `mapp1`

-   `mapp2`

```{r}
if (dfd == "twitter") {
  df_pop_dfd <- read.csv(paste0(wd, "/data/inputs/twitter/monthly/active_population.csv")) %>% 
    filter(time == "2021-03")
  } else if (dfd == "fb_tts_census_week") {
  df_pop_dfd <- read.csv(paste0(wd, "/data/inputs/fb/census-week/tts/active_population.csv")) %>% 
  rename("name" = "LAD21NM")
  } else if (dfd == "fb_tts_census_month") {
  df_pop_dfd <- read.csv(paste0(wd, "/data/inputs/fb/census-month/tts/active_population.csv")) %>% 
  rename("name" = "LAD21NM")
  } else if (dfd == "fb_stt_census_week") {
  df_pop_dfd <- read.csv(paste0(wd, "/data/inputs/fb/census-week/stt/active_population.csv")) %>% 
  rename("name" = "LAD21NM")
  } else if (dfd == "fb_stt_census_month") {
  df_pop_dfd <- read.csv(paste0(wd, "/data/inputs/fb/census-month/stt/active_population.csv")) %>% 
  rename("name" = "LAD21NM")
  } else if (dfd == "mapp1") {
  df_pop_dfd <- st_read(paste0(wd, "/data/inputs/mapp1/locations-home/lad/home-location-count_lda.gpkg")) %>% 
  rename("pop_dfd" = "home_location_count") %>% st_drop_geometry() %>% as.data.frame()
  } else if (dfd == "mapp2") {
  df_pop_dfd <- read.csv(paste0(wd, "/data/inputs/mapp2/locations-origin/lad/location-home-count_lad_htw.csv")) %>% 
  rename("code" = "LAD22CD")
  }
```

```{r}
sum(df_pop_dfd$pop_dfd)
```

# Measure bias

## Computing bias with available data

```{r}
if (dfd == "mapp1" | dfd == "mapp2") {
  df_bias <- df_pop_census %>% left_join(df_pop_dfd, by = "code")
} else {
  df_bias <- df_pop_census %>% left_join(df_pop_dfd, by = "name")
}

df_bias$bias <- 1 - df_bias$pop_dfd/df_bias$pop
df_bias$map_bias <- (1 - df_bias$pop_dfd/df_bias$pop)*100

```

## Imputing NA values of dfd pop according to linear regression and measuring missing values of bias

```{r}
# Compute standardised weights matrix based on boundaries for LADs
st_bias_map <- df_boundaries_hex %>%  filter(!grepl("^S", code))  %>% filter(!grepl("^N", code)) %>% filter(!grepl("^u", code))  %>% left_join(df_bias, by = "code") %>% st_as_sf()

model <- lm(pop_dfd ~ pop, data = st_bias_map, na.action = na.omit)

st_bias_map <- st_bias_map %>%
  mutate(pop_dfd = ifelse(is.na(pop_dfd), predict(model, newdata = .), pop_dfd))

st_bias_map <-  st_bias_map %>%
  mutate(bias = 1-pop_dfd/pop) %>%
  mutate(map_bias = bias*100)  

model <- lm(pop_dfd ~ pop, data = df_bias, na.action = na.omit)
df_bias <- df_bias %>%
  mutate(pop_dfd = ifelse(is.na(pop_dfd), predict(model, newdata = .), pop_dfd))
df_bias <-  df_bias %>%
  mutate(bias = 1-pop_dfd/pop) %>%
  mutate(map_bias = bias*100)
```

```{r}
if (dfd == "twitter"){
  write.csv(df_bias, paste0(wd, "/data/outputs/twitter/monthly/populations/active-population-bias/active_population_bias_lad.csv"))
  } else if (dfd == "fb_tts_census_week") {
  write.csv(df_bias, paste0(wd, "/data/outputs/fb/census-week/populations/tts/active-population-bias/active_population_bias_lad.csv"))
  } else if (dfd == "fb_tts_census_month") {
  write.csv(df_bias, paste0(wd, "/data/outputs/fb/census-month/populations/tts/active-population-bias/active_population_bias_lad.csv"))
  } else if (dfd == "fb_stt_census_week") {
  write.csv(df_bias, paste0(wd, "/data/outputs/fb/census-week/populations/stt/active-population-bias/active_population_bias_lad.csv"))
  } else if (dfd == "fb_stt_census_month") {
  write.csv(df_bias, paste0(wd, "/data/outputs/fb/census-month/populations/stt/active-population-bias/active_population_bias_lad.csv"))
  } else if (dfd == "mapp1") {
  write.csv(df_bias, paste0(wd, "/data/outputs/mapp1/census-month/populations/active-population-bias/active_population_bias_lad.csv"))
  } else if (dfd == "mapp2") {
  write.csv(df_bias, paste0(wd, "/data/outputs/mapp2/census-month/populations/active-population-bias/active_population_bias_lad.csv"))
  }

```

## Shared base-case model fit

```{r}
base_case_df <- df_bias %>%
  dplyr::select(code, name, pop, pop_dfd, bias) %>%
  filter(is.finite(pop), is.finite(pop_dfd), is.finite(bias))

# Proportional baseline used across sections: pop_dfd = alpha_hat * pop
base_case_model <- lm(pop_dfd ~ 0 + pop, data = base_case_df)
base_case_summary <- summary(base_case_model)
alpha_hat <- unname(coef(base_case_model)["pop"])
alpha_ci <- confint(base_case_model, "pop", level = 0.95)
alpha_ci_halfwidth <- (alpha_ci[2] - alpha_ci[1]) / 2
base_case_r2 <- as.numeric(base_case_summary$r.squared)
base_case_pred_bias <- 1 - alpha_hat
```

## Scatter plot

```{r}

# Compute the breaks using Jenks natural breaks
if (dfd == "twitter") { 
  num_classes <- 9  # Specify how many breaks you want
} else {
  num_classes <- 8
}

jenks_breaks <- classIntervals(df_bias$bias*100, n = num_classes, style = "jenks")$brks

jenks_breaks <- unique(jenks_breaks)

# Create a factor variable for color breaks
df_bias$jenks_bins <- cut(df_bias$bias*100, breaks = jenks_breaks, include.lowest = TRUE)

# Create a custom color palette from viridis
jenks_colors <- viridis(length(jenks_breaks) - 1)

# Manually format the legend labels to avoid scientific notation
formatted_labels <- sapply(1:(length(jenks_breaks) - 1), function(i) {
  paste0("[", 
         sprintf("%.1f", jenks_breaks[i]), ", ", 
         sprintf("%.1f", jenks_breaks[i + 1]), ")")
})


annotation <- data.frame(
   x = c(max(df_bias$pop*0.74, na.rm = TRUE), 
         max(df_bias$pop*0.74, na.rm = TRUE)),
   y = c(max(df_bias$pop_dfd *0.30, na.rm = TRUE), 
         max(df_bias$pop_dfd *0.18, na.rm = TRUE)),
   label = c(paste0("â = ", round(alpha_hat, 4),
                    " \u00B1 ", round(alpha_ci_halfwidth, 4)),
             paste0("R² = ", round(base_case_r2, 3)))
)

# Create the plot
ggplot(df_bias, aes(x = pop, y = pop_dfd, color = jenks_bins)) +
  geom_point(size = 5, alpha = 0.6, stroke = 0.7) +
  scale_color_viridis_d(
    na.translate = TRUE,
    na.value = "gray"  # Color for NA values
  ) +
  labs(
    x = "Population",
    y = "Population DT"
  ) +
  geom_abline(intercept = 0, slope = alpha_hat, linetype = "dotted", linewidth = 1.2, color = "black") +
  geom_text(data=annotation, aes(x=x, y=y, label=label),
           color="black", 
           size=23, 
           angle=0, 
           fontface="bold",
           family = "robotocondensed") +
  scale_x_continuous(labels = label_number(scale = 0.001, suffix = " k")) +
  scale_y_continuous(labels = label_number(scale = 0.001, suffix = " k")) +
  theme_plot_tufte() +
  theme(axis.title.x = element_text(size = 60),  # Set font size for X-axis title
    axis.title.y = element_text(size = 60),   # Set font size for Y-axis title
    axis.text.x = element_text(size = 50),     # Font size for X-axis ticks
    axis.text.y = element_text(size = 50),
    legend.position = "none",
    plot.margin = unit(c(1, 2, 1, 1), "lines")
    )

if (dfd == "twitter") {
  ggsave(paste0(wd_local, "/bias-detection/outputs/preliminary/twitter/size-bias/scatter/twitter_lad.png"), bg="white")
  } else if (dfd == "fb_tts_census_week") {
  ggsave(paste0(wd_local, "/bias-detection/outputs/preliminary/fb_tts/size-bias/scatter/fb_tts_lad_census_week.png"), bg="white")
  } else if (dfd == "fb_tts_census_month") {
  ggsave(paste0(wd_local, "/bias-detection/outputs/preliminary/fb_tts/size-bias/scatter/fb_tts_lad_census_month.png"), bg="white")
  } else if (dfd == "fb_stt_census_week") {
  ggsave(paste0(wd_local, "/bias-detection/outputs/preliminary/fb_stt/size-bias/scatter/fb_stt_lad_census_week.png"), bg="white")
  } else if (dfd == "fb_stt_census_month") {
  ggsave(paste0(wd_local, "/bias-detection/outputs/preliminary/fb_stt/size-bias/scatter/fb_stt_lad_census_month.png"), bg="white")
  } else if (dfd == "mapp1") {
  ggsave(paste0(wd_local, "/bias-detection/outputs/preliminary/mapp1/size-bias/scatter/mapp1_lad_census_month.png"), bg="white")
  } else if (dfd == "mapp2") {
  ggsave(paste0(wd_local, "/bias-detection/outputs/preliminary/mapp2/size-bias/scatter/mapp2_lad_census_month.png"), bg="white")
  }
```

## Histogram bias

```{r}

# Compute the breaks using Jenks natural breaks
if (dfd == "twitter") { 
  num_classes <- 9  # Specify how many breaks you want
} else {
  num_classes <- 8
}

jenks_breaks <- classIntervals(df_bias$bias*100, n = num_classes, style = "jenks")$brks

# Create a data frame with histogram information
hist_data <- data.frame(value = df_bias$bias*100)

jenks_breaks <- unique(jenks_breaks)

# Create a factor variable for color breaks
hist_data$jenks_bins <- cut(hist_data$value, breaks = jenks_breaks, include.lowest = TRUE)

# Create a custom color palette from viridis
jenks_colors <- viridis(length(jenks_breaks) - 1)

# Manually format the legend labels to avoid scientific notation
formatted_labels <- sapply(1:(length(jenks_breaks) - 1), function(i) {
  paste0("[", 
         sprintf("%.1f", jenks_breaks[i]), ", ", 
         sprintf("%.1f", jenks_breaks[i + 1]), ")")
})

if (dfd == "twitter") { 
  minimum_x <- sort(hist_data$value)[2] - 0.3
  binwidth <- .009
} else if (dfd != "mapp2") {
  minimum_x <- sort(hist_data$value)[1]
  binwidth <- .12
} else if (dfd == "mapp2") {
  minimum_x <- sort(hist_data$value)[1]
  binwidth <- .02
}

# Plot the histogram with ggplot
ggplot(hist_data, aes(x = value, fill = jenks_bins)) +
  xlim(minimum_x, 100) +
  geom_histogram(binwidth=binwidth, color = "black", linewidth = 0.3) +  # Adjust binwidth as needed
  scale_fill_viridis_d(
    name = "Size of bias",
    labels = formatted_labels,
    na.translate = TRUE,
    na.value = "gray"  # Set color for NA values
  ) +
  labs(x = "Size of bias", y = "Frequency") +
  theme_plot_tufte() +
  theme(
    legend.text = element_text(size = 60),       # Increase legend text size
    legend.title = element_text(size = 60),      # Increase legend title size
    legend.position = "right",                     # Optional: adjust legend position
    axis.title.x = element_text(size = 63),  # Set font size for X-axis title
    axis.title.y = element_text(size = 63),   # Set font size for Y-axis title
    axis.text.x = element_text(size = 50),     # Font size for X-axis ticks
    axis.text.y = element_text(size = 50)      # Font size for Y-axis ticks
  )

if (dfd == "twitter") {
  ggsave(paste0(wd_local, "/bias-detection/outputs/preliminary/twitter/size-bias/histogram/twitter_lad.png"), bg="white")
  } else if (dfd == "fb_tts_census_week") {
  ggsave(paste0(wd_local, "/bias-detection/outputs/preliminary/fb_tts/size-bias/histogram/fb_tts_lad_census_week.png"), bg="white")
  } else if (dfd == "fb_tts_census_month") {
  ggsave(paste0(wd_local, "/bias-detection/outputs/preliminary/fb_tts/size-bias/histogram/fb_tts_lad_census_month.png"), bg="white")
  } else if (dfd == "fb_stt_census_week") {
  ggsave(paste0(wd_local, "/bias-detection/outputs/preliminary/fb_stt/size-bias/histogram/fb_stt_lad_census_week.png"), bg="white")
  } else if (dfd == "fb_stt_census_month") {
  ggsave(paste0(wd_local, "/bias-detection/outputs/preliminary/fb_stt/size-bias/histogram/fb_stt_lad_census_month.png"), bg="white")
  } else if (dfd == "mapp1") {
  ggsave(paste0(wd_local, "/bias-detection/outputs/preliminary/mapp1/size-bias/histogram/mapp1_lad_census_month.png"), bg="white")
  } else if (dfd == "mapp2") {
  ggsave(paste0(wd_local, "/bias-detection/outputs/preliminary/mapp2/size-bias/histogram/mapp2_lad_census_month.png"), bg="white")
  }
```

# Spatial autocorrelation

```{r}
st_bias_map <- df_boundaries %>% left_join(df_bias, by = "code") %>% st_as_sf() %>% filter(!grepl("^S", code))
```

## Spatial weights matrix

Choose between:

-   `'queen'` (queen neighbourhood)

-   `'fbw'` (fixed band-width, weights as 1/d)

-   `'knn'` (k nearest-neighbours, weights as 1/d)

-   `'db'` (distance band, weights as 1/d)

### Option 1: `queen`

```{r}
if (w_scheme == 'queen'){
  if ("bias_w" %in% colnames(st_bias_map)) {
    st_bias_map <- st_bias_map[, !names(st_bias_map) %in% "bias_w"]
  }
  suffix <- '_w_queen'
  # Extract the coordinates of the centroids (or your points)
  coords <- st_coordinates(st_centroid(st_bias_map))
  # Create queen neighbours list
  nb_q <- poly2nb(st_bias_map, queen = TRUE)
  # Create a row-standardised spatial weights matrix using distance-based neighbors
  w_matrix <- nb2listw(nb_q, style = "W", zero.policy=TRUE)
  # Compute spatial lag
  bias_w <- lag.listw(w_matrix, st_bias_map$bias, NAOK=TRUE)
  # Save spatial lag in additional column
  st_bias_map$bias_w <- bias_w
}
```

### Option 2: finding optimal fixed band width using `gwr.sel()` function from `spgwr` package

```{r}

if (w_scheme == 'fbw'){
  if ("bias_w" %in% colnames(st_bias_map)) {
    st_bias_map <- st_bias_map[, !names(st_bias_map) %in% "bias_w"]
  }
  suffix <- '_w_fbw'
  coords <- st_coordinates(st_centroid(st_bias_map))
  # find optimal kernel bandwidth using cross validation
  fbw <- gwr.sel(map_bias ~ 1, 
                 data = st_bias_map, 
                 coords=cbind( coords[, "X"], coords[, "Y"]),
                 longlat = TRUE,
                 adapt = FALSE, 
                 gweight = gwr.Gauss, 
                 verbose = FALSE)
  # Create a distance-based neighbors list with a minimum distance of 0 and maximum distance given by fbw
  nb_d <- dnearneigh(st_coordinates(st_centroid(st_bias_map)), d1=0, d2=fbw)
  # Create a row-standardised spatial weights matrix using distance-based neighbors
  w_matrix <- nb2listw(nb_d, style = "W", zero.policy=TRUE)
  # WARNING!!!!! This weights matrix leaves too many observations disconnected. Below we print the percentage of observations with no neighbors:
  # Find the number of observations with zero neighbors
  zero_neighbors_count <- sum(unlist(lapply(w_matrix$weights, length)) == 0)
  zero_neighbors_count/nrow(st_bias_map)*100
  # Compute spatial lag
  bias_w <- lag.listw(w_matrix, st_bias_map$bias, NAOK=TRUE)
  # Save spatial lag in additional column
  st_bias_map$bias_w <- bias_w
}
```

Too many.

### Option 3: neighbourhoods based on k-nearest neighbours, weights as inverse of distance

```{r}

if (w_scheme == 'knn'){
  if ("bias_w" %in% colnames(st_bias_map)) {
    st_bias_map <- st_bias_map[, !names(st_bias_map) %in% "bias_w"]
  }
  suffix <- '_w_knn'
  # Extract the coordinates of the centroids (or your points)
  coords <- st_coordinates(st_centroid(st_bias_map))
  # Create k-Nearest Neighbors list with k = 10 (using centroids of a spatial object)
  nb_knn <- knearneigh(coords, k = 10)
  # Create weights matrix with equal weights
  w_knn <- knn2nb(nb_knn)
  # Compute distances from observation to neighbours
  k.distances <- nbdists(w_knn, coords)
  # Use these distances as weights
  invd2a <- lapply(k.distances, function(x) (1/(x/100)))
  # Create weights matrix with weights as 1/d, normalised
  w_matrix <- nb2listw(w_knn, glist = invd2a, style = "W")
  # Compute spatial lag
  bias_w <- lag.listw(w_matrix, st_bias_map$bias, NAOK=TRUE)
  # Save spatial lag in additional column
  st_bias_map$bias_w <- bias_w
}  

```

### Option 4: neighbourhoods based on distance band (set by user, unlike fbw which is "optimally" computed), weights as inverse of distance

```{r}

if (w_scheme == 'db'){
  if ("bias_w" %in% colnames(st_bias_map)) {
    st_bias_map <- st_bias_map[, !names(st_bias_map) %in% "bias_w"]
  }
  suffix <- '_w_db'
  # Extract the coordinates of the centroids (or your points)
  coords <- st_coordinates(st_centroid(st_bias_map))
  # Compute distance band to ensure that all the observations have neighbours
  k1 <- knn2nb(knearneigh(coords))
  critical.threshold <- max(unlist(nbdists(k1, coords)))
  # Compute lists of neighbours 
  nb.dist.band <- dnearneigh(coords, 0, critical.threshold)
  # Compute distances
  distances <- nbdists(nb.dist.band, coords)
  # Compute inverse distances
  invd1a <- lapply(distances, function(x) (1/(x/100)))
  # Compute spatial weights matrix, normalised
  w_matrix <- nb2listw(nb.dist.band, glist = invd1a, style = "W", zero.policy=TRUE)
  # Compute spatial lag
  bias_w <- lag.listw(w_matrix, st_bias_map$bias, NAOK=TRUE)
  # Save spatial lag in additional column
  st_bias_map$bias_w <- bias_w
}  
```

### Save data with spatial lag

```{r}
if (dfd == "twitter"){
  write.csv(st_drop_geometry(st_bias_map), 
            paste0(wd, "/data/outputs/twitter/monthly/populations/active-population-bias/active_population_bias_lad", suffix, ".csv"))
} else if (dfd == "fb_tts_census_week") {
  write.csv(st_drop_geometry(st_bias_map), 
            paste0(wd, "/data/outputs/fb/census-week/populations/tts/active-population-bias/active_population_bias_lad", suffix, ".csv"))
} else if (dfd == "fb_tts_census_month") {
  write.csv(st_drop_geometry(st_bias_map), 
            paste0(wd, "/data/outputs/fb/census-month/populations/tts/active-population-bias/active_population_bias_lad", suffix, ".csv"))
} else if (dfd == "fb_stt_census_week") {
  write.csv(st_drop_geometry(st_bias_map), 
            paste0(wd, "/data/outputs/fb/census-week/populations/stt/active-population-bias/active_population_bias_lad", suffix, ".csv"))
} else if (dfd == "fb_stt_census_month") {
  write.csv(st_drop_geometry(st_bias_map), 
            paste0(wd, "/data/outputs/fb/census-month/populations/stt/active-population-bias/active_population_bias_lad", suffix, ".csv"))
} else if (dfd == "mapp1") {
  write.csv(st_drop_geometry(st_bias_map), 
            paste0(wd, "/data/outputs/mapp1/census-month/populations/active-population-bias/active_population_bias_lad", suffix, ".csv"))
} else if (dfd == "mapp2") {
  write.csv(st_drop_geometry(st_bias_map), 
            paste0(wd, "/data/outputs/mapp2/census-month/populations/active-population-bias/active_population_bias_lad", suffix, ".csv"))
}
```

## Moran's I

```{r}
# Moran's I Monte Carlo test
result <- moran.mc(st_bias_map$bias, w_matrix, nsim=1000, alternative="two.sided")

# Extract Moran's I and p-value
mI <- as.numeric(result$statistic)  # Moran's I statistic
p_value_mI <- as.numeric(result$p.value)    # p-value
```

```{r}
mI
p_value_mI
```

# Linear baseline

## Fit linear validation model and report $R^2$

```{r}

base_case_rmse <- sqrt(mean((base_case_df$bias - base_case_pred_bias)^2))

base_case_metrics <- data.frame(
  dataset = dfd,
  n_lad = nrow(base_case_df),
  alpha_hat = alpha_hat,
  predicted_bias = base_case_pred_bias,
  r_squared = base_case_r2,
  rmse = base_case_rmse
)

write.csv(
  base_case_metrics,
  file.path(output_data_dir, "base-case", paste0("base_case_metrics", suffix, ".csv")),
  row.names = FALSE
)

base_case_metrics
```

## Map residuals from the base-case model

```{r}

st_bias_map <- df_boundaries_hex %>%  filter(!grepl("^S", code))  %>% filter(!grepl("^N", code)) %>% filter(!grepl("^u", code))  %>% left_join(df_bias, by = "code") %>% st_as_sf()

st_base_case_map <- st_bias_map %>%
  mutate(
    base_case_pred = base_case_pred_bias,
    base_case_resid = (bias - base_case_pred)/base_case_pred*100
  )

# For Twitter, City of London is an extreme residual; set to NA in map display only.
if (dfd == "twitter") {
  st_base_case_map <- st_base_case_map %>%
    mutate(base_case_resid_map = ifelse(code == "E09000001", NA_real_, base_case_resid))
} else {
  st_base_case_map <- st_base_case_map %>%
    mutate(base_case_resid_map = base_case_resid)
}

# Standardise residuals to [-1, 1] for colouring, excluding NA (Twitter's City of London).
resid_pos_max <- max(st_base_case_map$base_case_resid_map[st_base_case_map$base_case_resid_map > 0], na.rm = TRUE)
resid_neg_min <- min(st_base_case_map$base_case_resid_map[st_base_case_map$base_case_resid_map < 0], na.rm = TRUE)

st_base_case_map <- st_base_case_map %>%
  mutate(
    base_case_resid_plot = dplyr::case_when(
      is.na(base_case_resid_map) ~ NA_real_,
      base_case_resid_map >= 0 & is.finite(resid_pos_max) & resid_pos_max > 0 ~ pmin(base_case_resid_map / resid_pos_max, 1),
      base_case_resid_map < 0 & is.finite(resid_neg_min) & resid_neg_min < 0 ~ pmax(base_case_resid_map / abs(resid_neg_min), -1),
      TRUE ~ 0
    )
  )

# Moran's I on residual bias (computed here for map annotation)
resid_valid <- is.finite(st_base_case_map$base_case_resid)
if (!all(resid_valid)) {
  stop("Non-finite residuals found in base-case model. Moran's I on residuals requires finite values.")
}

resid_moran <- moran.mc(
  st_base_case_map$base_case_resid,
  w_matrix,
  nsim = 1000,
  alternative = "two.sided",
  zero.policy = TRUE
)

resid_mI <- as.numeric(resid_moran$statistic)
resid_p_value_mI <- as.numeric(resid_moran$p.value)

base_case_spatial <- data.frame(
  dataset = dfd,
  moran_i_residuals = resid_mI,
  p_value_residuals = resid_p_value_mI
)

p_annotation_resid <- ifelse(resid_p_value_mI < 0.05, "p < 0.05", paste("p =", round(resid_p_value_mI, 2)))
bbox <- st_bbox(st_base_case_map)
xmin <- bbox["xmin"]
xmax <- bbox["xmax"]
ymin <- bbox["ymin"]
ymax <- bbox["ymax"]
annotation_resid <- data.frame(
   x = c(xmin + 0.1*(xmax-xmin), xmin + 0.1*(xmax-xmin)),
   y = c(ymin + 0.85*(ymax-ymin), ymin + 0.78*(ymax-ymin)),
   label = c(paste0("I = ", round(resid_mI,2)), p_annotation_resid)
)

residual_map <- ggplot(st_base_case_map) +
  geom_sf(aes(fill = base_case_resid_plot)) +
  scale_fill_gradientn(
    colours = c("#2166AC", "white", "#B2182B"),
    values = scales::rescale(c(-1, 0, 1), from = c(-1, 1)),
    limits = c(-1, 1),
    na.value = "#2166AC",
    name = "Bias residual (%)\n[standardised -1 to 1]"
  ) +
  labs(title = " ") +
  geom_text(data=annotation_resid, aes(x=x, y=y, label=label),
           color="black",
           size=30,
           angle=0,
           fontface="bold",
           family = "robotocondensed") +
  theme_map_tufte() +
  theme(
    legend.text = element_text(size = 60),
    legend.title = element_text(size = 60,
                                face = "bold", 
                                lineheight = 0.3, 
                                margin = margin(b = 12)),
    legend.position = "right",
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank()
  )

ggsave(
  filename = file.path(output_fig_dir, "base-case", paste0(dfd, "_lad_base_case_residual_map_hex", suffix, ".png")),
  plot = residual_map,
  bg = "white",
  width = 10,
  height = 8,
  units = "in"
)
```

```{r}
residual_map
```

```{r}
write.csv(
  base_case_spatial,
  file.path(output_data_dir, "base-case", paste0("base_case_residual_moran", suffix, ".csv")),
  row.names = FALSE
)

base_case_spatial
```

## Demographic associations of base-case residuals

```{r}
df_covariates <- read.csv(paste0(wd, "/data/inputs/census/census2021-ts/combined-data-ltla.csv"))
covariate_vars <- names(df_covariates)[sapply(df_covariates, is.numeric)]
covariate_vars <- setdiff(
  covariate_vars,
  c("date", "geography.code", "geography", "code", "name")
)

st_base_case_map_clean <- st_base_case_map  %>%
  dplyr::filter(code != "E09000001")


resid_demog <- st_drop_geometry(st_base_case_map_clean) %>%
  dplyr::select(code, pop, base_case_resid, base_case_resid_plot) %>%
  left_join(df_covariates, by = join_by(code == geography.code))

run_pairwise_cor <- function(vname) {
  test_df <- resid_demog %>%
    dplyr::select(base_case_resid, value = all_of(vname)) %>%
    filter(is.finite(base_case_resid), is.finite(value))

  if (nrow(test_df) < 30 || stats::sd(test_df$value) == 0) {
    return(NULL)
  }

  gam_linear <- mgcv::gam(base_case_resid ~ value, data = test_df, method = "REML")
  gam_smooth <- mgcv::gam(base_case_resid ~ s(value, k = 5), data = test_df, method = "REML")
  lm_model <- stats::lm(base_case_resid ~ value, data = test_df)
  pearson_test <- suppressWarnings(cor.test(test_df$base_case_resid, test_df$value, method = "pearson"))
  spearman_test <- suppressWarnings(cor.test(test_df$base_case_resid, test_df$value, method = "spearman", exact = FALSE))
  
  if (requireNamespace("energy", quietly = TRUE)) {
    dcor_value <- tryCatch(
      energy::dcor(test_df$base_case_resid, test_df$value),
      error = function(e) NA_real_
    )
  } else {
    dcor_value <- NA_real_
  }

  data.frame(
    dataset = dfd,
    variable = vname,
    n = nrow(test_df),
    pearson_r = unname(pearson_test$estimate),
    pearson_p = pearson_test$p.value,
    spearman_rho = unname(spearman_test$estimate),
    spearman_p = spearman_test$p.value,
    distance_correlation = dcor_value,
    lm_aic = AIC(lm_model),
    lm_bic = BIC(lm_model),
    gam_aic = AIC(gam_smooth),
    gam_bic = BIC(gam_smooth)
  )
}

correlation_tests <- lapply(covariate_vars, run_pairwise_cor) %>%
  bind_rows()

if (nrow(correlation_tests) == 0) {
  correlation_tests <- data.frame(
    dataset = character(0),
    variable = character(0),
    n = integer(0),
    pearson_r = numeric(0),
    pearson_p = numeric(0),
    spearman_rho = numeric(0),
    spearman_p = numeric(0),
    distance_correlation = numeric(0),
    lm_aic = numeric(0),
    lm_bic = numeric(0),
    gam_aic = numeric(0),
    gam_bic = numeric(0)
  )
} else {
  correlation_tests <- correlation_tests %>%
    mutate(
      delta_aic = lm_aic - gam_aic,
      delta_bic = lm_bic - gam_bic
    ) %>%
    arrange(desc(delta_aic))
}

# Select demographic variable with the lowest GAM AIC for follow-up plotting
df_dictionary <- read.csv(paste0(wd, "/data/inputs/census/census2021-ts/dictionary.csv"))
best_demog_var <- correlation_tests %>%
  dplyr::slice_min(order_by = gam_aic, n = 1, with_ties = FALSE) %>%
  dplyr::pull(variable)

if (length(best_demog_var) == 0) {
  stop("No demographic variable from dictionary found in correlation_tests.")
}
best_demog_var <- best_demog_var[[1]]
x_label <- df_dictionary %>%
  dplyr::filter(var_label5 == best_demog_var) %>%
  dplyr::slice(1) %>%
  dplyr::pull(var_label2)
if (length(x_label) == 0 || is.na(x_label)) {
  x_label <- best_demog_var
}

write.csv(
  correlation_tests,
  file.path(output_data_dir, "base-case", paste0("base_case_residual_pairwise_correlations", suffix, ".csv")),
  row.names = FALSE
)

correlation_tests
```

```{r}
best_demog_var

resid_demog_clean <- resid_demog %>%
  dplyr::filter(is.finite(base_case_resid_plot), is.finite(.data[[best_demog_var]]), is.finite(pop)) %>%
  dplyr::filter(code != "E09000001")

y_abs_max <- max(abs(resid_demog_clean$base_case_resid), na.rm = TRUE)

plot_demog_assoc <- ggplot(resid_demog_clean, aes(x = .data[[best_demog_var]], y = base_case_resid)) +
  geom_point(aes(size = pop, fill = base_case_resid_plot), alpha = 0.8, shape = 21, color = "black", stroke = 0.25) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.2, color = "black", linetype = "dotted") +
  geom_smooth(method = "gam", formula = y ~ s(x, k = 5), se = FALSE, linewidth = 1.2, color = "#B2182B") +
  scale_y_continuous(limits = c(-y_abs_max, y_abs_max)) +
  labs(
    x = x_label,
    y = "Residual bias (%)"
  ) +
  scale_fill_gradientn(
    colours = c("#2166AC", "white", "#B2182B"),
    values = scales::rescale(c(-1, 0, 1), from = c(-1, 1)),
    limits = c(-1, 1),
    na.value = "#2166AC",
    name = "Bias residual (%)\n[standardised -1 to 1]"
  ) +
  scale_size_continuous(name = "Population", labels = label_number(scale = 0.001, suffix = " k")) +
  theme_plot_tufte() +
  theme(
    axis.title.x = element_text(size = 85, face = "bold"),
    axis.title.y = element_text(size = 85, face = "bold"),
    axis.text.x = element_text(size = 87, face = "bold"),
    axis.text.y = element_text(size = 87, face = "bold"),
    legend.title = element_text(size = 87, face = "bold", lineheight = 0.3, margin = margin(b = 12)),
    legend.text = element_text(size = 81),
    plot.margin = unit(c(1, 2, 1, 1), "lines")
  )

ggsave(
  filename = file.path(output_fig_dir, "base-case", paste0(dfd, "_residual_demographic_association_gam", suffix, ".png")),
  plot = plot_demog_assoc,
  bg = "white",
  width = 10,
  height = 7,
  units = "in"
)

plot_demog_assoc
```

## Save base-case residual table

```{r}
base_case_residuals <- st_drop_geometry(st_base_case_map) %>%
  dplyr::select(code, name, pop, pop_dfd, bias, base_case_pred, base_case_resid)

write.csv(
  base_case_residuals,
  file.path(output_data_dir, "base-case", paste0("base_case_residuals", suffix, ".csv")),
  row.names = FALSE
)
```

# Map bias

## England and Wales

### Static map

```{r}
# # Replace the specific entry with NA
# # st_bias_map[st_bias_map$code == "E09000001", "map_bias"] <- NA
# 
# # Create a new factor based on Jenks breaks
# st_bias_map$jenks_bins <- cut(st_bias_map$map_bias, jenks_breaks, include.lowest = TRUE)
# 
# 
# ggplot(data = st_bias_map) +
#   geom_sf(aes(fill = jenks_bins)) +
#   scale_fill_viridis_d(name = "Size of bias",
#     labels = formatted_labels,
#     na.translate = TRUE,
#     na.value = "gray") +  # Set color for NA values
#     labs(title = " ") +
#     theme_map_tufte() +
#     theme(
#       legend.text = element_text(size = 60),
#       legend.title = element_text(size = 60),
#       legend.position = "right"
#       )
# 
# if (dfd == "twitter") {
#   ggsave(paste0(wd_local, "/bias-detection/outputs/preliminary/twitter/size-bias/map/twitter_lad.png"), bg="white")
#   } else if (dfd == "fb_tts_census_week") {
#   ggsave(paste0(wd_local, "/bias-detection/outputs/preliminary/fb_tts/size-bias/map/fb_tts_lad_census_week.png"), bg="white")
#   } else if (dfd == "fb_tts_census_month") {
#   ggsave(paste0(wd_local, "/bias-detection/outputs/preliminary/fb_tts/size-bias/map/fb_tts_lad_census_month.png"), bg="white")
#   } else if (dfd == "fb_stt_census_week") {
#   ggsave(paste0(wd_local, "/bias-detection/outputs/preliminary/fb_stt/size-bias/map/fb_stt_lad_census_week.png"), bg="white")
#   } else if (dfd == "fb_stt_census_month") {
#   ggsave(paste0(wd_local, "/bias-detection/outputs/preliminary/fb_stt/size-bias/map/fb_stt_lad_census_month.png"), bg="white")
#   }
```

### Hexagons

```{r}
df_bias 

st_bias_map <- df_boundaries_hex %>%  filter(!grepl("^S", code))  %>% filter(!grepl("^N", code)) %>% filter(!grepl("^u", code))  %>% left_join(df_bias, by = "code") %>% st_as_sf()

# Replace the specific entry with NA
# st_bias_map[st_bias_map$code == "E09000001", "map_bias"] <- NA

# Create a new factor based on Jenks breaks
st_bias_map$jenks_bins <- cut(st_bias_map$map_bias, jenks_breaks, include.lowest = TRUE)

p_annotation <- ifelse(p_value_mI < 0.05, "p < 0.05", paste("p =", round(p_value_mI, 2)))
bbox <- st_bbox(st_bias_map)  # Get the map's bounding box
xmin <- bbox["xmin"]
xmax <- bbox["xmax"]
ymin <- bbox["ymin"]
ymax <- bbox["ymax"]
annotation <- data.frame(
   x = c(xmin + 0.1*(xmax-xmin), xmin + 0.1*(xmax-xmin)),
   y = c(ymin + 0.85*(ymax-ymin), ymin + 0.75*(ymax-ymin)),
   label = c(paste0("I = ", round(mI,2)), p_annotation)
)

ggplot(data = st_bias_map) +
  geom_sf(aes(fill = jenks_bins)) +
  scale_fill_viridis_d(name = "Size of bias",
    labels = formatted_labels,
    na.translate = FALSE,
    na.value = "gray") +  # Set color for NA values
    labs(title = " ") +
    geom_text(data=annotation, aes(x=x, y=y, label=label),
           color="black",
           size=23,
           angle=0,
           fontface="bold",
           family = "robotocondensed") +
    theme_map_tufte() +
    theme(
      legend.text = element_text(size = 55),
      legend.title = element_text(size = 55),
      legend.position = "right",
      axis.title = element_blank(),        # Remove axis titles
      axis.text = element_blank(),         # Remove axis text
      axis.ticks = element_blank()         # Remove axis ticks
      )

if (dfd == "twitter") {
  ggsave(paste0(wd_local, "/bias-detection/outputs/preliminary/twitter/size-bias/map/twitter_lad_hex", suffix, ".png"), bg="white")
  } else if (dfd == "fb_tts_census_week") {
  ggsave(paste0(wd_local, "/bias-detection/outputs/preliminary/fb_tts/size-bias/map/fb_tts_lad_census_week_hex", suffix, ".png"), bg="white")
  } else if (dfd == "fb_tts_census_month") {
  ggsave(paste0(wd_local, "/bias-detection/outputs/preliminary/fb_tts/size-bias/map/fb_tts_lad_census_month_hex", suffix, ".png"), bg="white")
  } else if (dfd == "fb_stt_census_week") {
  ggsave(paste0(wd_local, "/bias-detection/outputs/preliminary/fb_stt/size-bias/map/fb_stt_lad_census_week_hex", suffix, ".png"), bg="white")
  } else if (dfd == "fb_stt_census_month") {
  ggsave(paste0(wd_local, "/bias-detection/outputs/preliminary/fb_stt/size-bias/map/fb_stt_lad_census_month_hex", suffix, ".png"), bg="white")
  } else if (dfd == "mapp1") {
  ggsave(paste0(wd_local, "/bias-detection/outputs/preliminary/mapp1/size-bias/map/mapp1_lad_census_month_hex", suffix, ".png"), bg="white")
  } else if (dfd == "mapp2") {
  ggsave(paste0(wd_local, "/bias-detection/outputs/preliminary/mapp2/size-bias/map/mapp2_lad_census_month_hex", suffix, ".png"), bg="white")
  }
```

### Interactive map

```{r}

# st_bias_map <- df_boundaries %>% left_join(df_bias, by = "code") %>% st_as_sf() %>% filter(!grepl("^S", code))
# 
# # Replace the specific entry with NA
# # st_bias_map[st_bias_map$code == "E09000001", "map_bias"] <- NA
# 
# # Create a new factor based on Jenks breaks
# st_bias_map$jenks_bins <- cut(st_bias_map$map_bias, jenks_breaks, include.lowest = TRUE)
# 
# tmap_mode("view") # enable interactivity
# interactive_map <- tm_shape(st_bias_map) + # input data
#     tm_borders() +
#     tm_fill("jenks_bins", # draw and fill polygons
#           palette = "viridis",
#           title = "Size of bias",
#           id="name")
# 
# if (dfd == "twitter"){
#   tmap_save(interactive_map, paste0(wd, "/bias-detection/outputs/preliminary/twitter/size-bias/interactive/twitter_lad.html"))
#   } else if (dfd == "fb_tts_census_week") {
#   tmap_save(interactive, paste0(wd, "/bias-detection/outputs/preliminary/fb_tts/size-bias/interactive/fb_tts_lad_census_week.html"))
#   } else if (dfd == "fb_tts_census_month") {
#   tmap_save(interactive, paste0(wd, "/bias-detection/outputs/preliminary/fb_tts/size-bias/interactive/fb_tts_lad_census_month.html"))
#   } else if (dfd == "fb_stt_census_week") {
#   tmap_save(interactive, paste0(wd, "/bias-detection/outputs/preliminary/fb_stt/size-bias/interactive/fb_stt_lad_census_week.html"))
#   } else if (dfd == "fb_stt_census_month") {
#   tmap_save(interactive, paste0(wd, "/bias-detection/outputs/preliminary/fb_stt/size-bias/interactive/fb_stt_lad_census_month.html"))
#   }
# 
# interactive_map
```

## London

### Static map

```{r}
# 
# st_bias_map <- df_boundaries %>% left_join(df_bias, by = "code") %>% st_as_sf() %>% filter(grepl("^E09", code)) 
# 
# st_bias_map[st_bias_map$code == "E09000001", "map_bias"] <- NA
# 
# st_bias_map$jenks <- cut(st_bias_map$map_bias, breaks$brks, include.lowest = TRUE)
# 
# ggplot(data = st_bias_map) +
#   geom_sf(aes(fill = jenks)) +
#   scale_fill_viridis_d(name = "Size of bias") +  # Adjust color palette
#   labs(title = " ") +
#   theme_map() +
#   theme(
#     legend.text = element_text(size = 24),       # Increase legend text size
#     legend.title = element_text(size = 30),      # Increase legend title size
#     legend.position = "right"                    # Optional: adjust legend position
#   )
# 
# if (dfd == "twitter"){
#   ggsave(paste0(wd_local, "/bias-detection/outputs/preliminary/twitter/size-bias/map/twitter_lad_london.png"), bg="white")
#   } else if (dfd == "fb_tts_census_week") {
#   ggsave(paste0(wd_local, "/bias-detection/outputs/preliminary/fb_tts/size-bias/map/fb_tts_lad_london_census_week.png"), bg="white")
#   } else if (dfd == "fb_tts_census_month") {
#   ggsave(paste0(wd_local, "/bias-detection/outputs/preliminary/fb_tts/size-bias/map/fb_tts_lad_london_census_month.png"), bg="white")
#   } else if (dfd == "fb_stt_census_week") {
#   ggsave(paste0(wd_local, "/bias-detection/outputs/preliminary/fb_stt/size-bias/map/fb_stt_lad_london_census_week.png"), bg="white")
#   } else if (dfd == "fb_stt_census_month") {
#   ggsave(paste0(wd_local, "/bias-detection/outputs/preliminary/fb_stt/size-bias/map/fb_stt_lad_london_census_month.png"), bg="white")
#   }
```

### Interactive map

```{r}
# tmap_mode("view") # enable interactivity
# interactive_map <- tm_shape(st_bias_map) + # input data
#     tm_borders() +
#     tm_fill("map_bias", # draw and fill polygons
#           palette = "viridis",
#           title = "Size of bias",
#           id="name")
# 
# if (dfd == "twitter"){
#   tmap_save(interactive_map, paste0(wd, "/bias-detection/outputs/preliminary/twitter/size-bias/interactive/twitter_lad_london.html"))
#   } else if (dfd == "fb_tts_census_week") {
#   tmap_save(interactive_map, paste0(wd, "/bias-detection/outputs/preliminary/fb_tts/size-bias/interactive/fb_tts_lad_london_census_week.html"))
#   } else if (dfd == "fb_tts_census_month") {
#   tmap_save(interactive_map, paste0(wd, "/bias-detection/outputs/preliminary/fb_tts/size-bias/interactive/fb_tts_lad_london_census_month.html"))
#   } else if (dfd == "fb_stt_census_week") {
#   tmap_save(interactive_map, paste0(wd, "/bias-detection/outputs/preliminary/fb_stt/size-bias/interactive/fb_stt_lad_london_census_week.html"))
#   } else if (dfd == "fb_stt_census_month") {
#   tmap_save(interactive_map, paste0(wd, "/bias-detection/outputs/preliminary/fb_stt/size-bias/interactive/fb_stt_lad_london_census_month.html"))
#   }
# 
# interactive_map
```
