---
title: "measure-bias"
format: html
editor: visual
---

```{r}
library(dplyr)
library(sf)
library(ggplot2)
library(tmap)
library(viridis)

library(sysfonts)
library(showtextdb)

library(classInt)

library(scales)

#| include = FALSE
source("./style/data-visualisation_theme.R")
```

# Set themes

Set font style

```{r}
# clean workspace
rm(list=ls())
# load font
font_add_google("Roboto Condensed", "robotocondensed")
# automatically use showtext to render text
showtext_auto()
```

Theme for maps

```{r}
theme_map <- function(...) {
  theme_tufte() +
  theme(
    text = element_text(family = "robotocondensed"),
    # remove all axes
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks = element_blank()
    )
}
```

Theme for plots

```{r}
theme_tufte2 <- function(...) {
  theme_tufte() +
  theme(
    text = element_text(family = "robotocondensed"),
    )
}

```

# Data import

## Census and boundaries

```{r}
wd <- "/Volumes/DEBIAS"
wd_local <- "/Users/carmen/Documents/github/de-bias"
```

```{r}
df_pop_census <- read.csv(paste0(wd, "/data/inputs/census/census2021-ts/census2021-ts001/census2021-ts001-ltla.csv")) %>% select(date, geography, geography.code, Residence.type..Total..measures..Value) %>% rename("code" = "geography.code", "name" = "geography", "pop" = "Residence.type..Total..measures..Value")
```

```{r}
# ONLY UNCOMMENT THE FIRST TIME YOU READ BOUNDARIES. FOR SUBSEQUENT TIMES, RUN CELL BELOW FOR SIMPLIFIED BOUNDARY FILE

# df_boundaries <- st_read(paste0(wd, "/data/inputs/geographies/boundaries/LAD_Dec_2021_GB_BFC_2022.gpkg")) %>% select(LAD21CD, SHAPE) %>% rename("code" = "LAD21CD") %>% st_simplify(preserveTopology = FALSE, dTolerance = 5000)
# 
# # Save as GeoPackage (GPKG)
# st_write(df_boundaries, paste0(wd, "/data/inputs/geographies/boundaries/LAD_Dec_2021_GB_BFC_2022_simplified.gpkg"))
```

```{r}
df_boundaries <- st_read(paste0(wd, "/data/inputs/geographies/boundaries/LAD_Dec_2021_GB_BFC_2022_simplified.gpkg"))
```

## Digital trace data

This first line defines the data source to be analysed. We hav1e five options:

-   `twitter`

-   `fb_tts_census_week`

-   `fb_tts_census_month`

-   `fb_stt_census_week`

-   `fb_stt_census_month`

```{r}
dfd <- "fb_stt_census_month"
```

```{r}
if (dfd == "twitter") {
  df_pop_dfd <- read.csv(paste0(wd, "/data/inputs/twitter/monthly/active_population.csv")) %>% 
    filter(time == "2021-03")
  } else if (dfd == "fb_tts_census_week") {
  df_pop_dfd <- read.csv(paste0(wd, "/data/inputs/fb/census-week/tts/active_population.csv")) %>% 
  rename("name" = "LAD21NM")
  } else if (dfd == "fb_tts_census_month") {
  df_pop_dfd <- read.csv(paste0(wd, "/data/inputs/fb/census-month/tts/active_population.csv")) %>% 
  rename("name" = "LAD21NM")
  } else if (dfd == "fb_stt_census_week") {
  df_pop_dfd <- read.csv(paste0(wd, "/data/inputs/fb/census-week/stt/active_population.csv")) %>% 
  rename("name" = "LAD21NM")
  } else if (dfd == "fb_stt_census_month") {
  df_pop_dfd <- read.csv(paste0(wd, "/data/inputs/fb/census-month/stt/active_population.csv")) %>% 
  rename("name" = "LAD21NM")
  } 
```

# Measure bias

```{r}
df_bias <- df_pop_census %>% left_join(df_pop_dfd, by = "name")
df_bias$bias <- 1 - df_bias$pop_dfd/df_bias$pop
df_bias$map_bias <- 1 - df_bias$pop_dfd/df_bias$pop
```

```{r}
if (dfd == "twitter"){
  write.csv(df_bias, paste0(wd, "/data/outputs/twitter/monthly/populations/active-population-bias/active_population_bias_lad.csv"))
  } else if (dfd == "fb_tts_census_week") {
  write.csv(df_bias, paste0(wd, "/data/outputs/fb/census-week/populations/tts/active-population-bias/active_population_bias_lad.csv"))
  } else if (dfd == "fb_tts_census_montbh") {
  write.csv(df_bias, paste0(wd, "/data/outputs/fb/census-month/populations/tts/active-population-bias/active_population_bias_lad.csv"))
  } else if (dfd == "fb_stt_census_week") {
  write.csv(df_bias, paste0(wd, "/data/outputs/fb/census-week/populations/stt/active-population-bias/active_population_bias_lad.csv"))
  } else if (dfd == "fb_stt_census_month") {
  write.csv(df_bias, paste0(wd, "/data/outputs/fb/census-month/populations/stt/active-population-bias/active_population_bias_lad.csv"))
  }

```

```{r}
plot(df_bias$pop, df_bias$bias)
```

## Histogram populations

```{r}
# Compute the breaks using Jenks natural breaks
num_classes <- 8  # Specify how many breaks you want
jenks_breaks <- classIntervals(df_bias$pop, n = num_classes, style = "jenks")$brks

# Create a data frame with histogram information
hist_data <- data.frame(value = df_bias$pop)

# Create a factor variable for color breaks
hist_data$jenks_bins <- cut(hist_data$value, breaks = jenks_breaks, include.lowest = TRUE)

# Create a custom color palette from viridis
jenks_colors <- viridis(length(jenks_breaks) - 1)

# Manually format the legend labels to avoid scientific notation
formatted_labels <- sapply(1:(length(jenks_breaks) - 1), function(i) {
  paste0("[", format(jenks_breaks[i], scientific = FALSE), ", ", 
         format(jenks_breaks[i + 1], scientific = FALSE), ")")
})

# Plot the histogram with ggplot
ggplot(hist_data, aes(x = value, fill = jenks_bins)) +
  geom_histogram(color = "black", size = 0.3, bins=60) +  # Adjust binwidth as needed
  scale_fill_manual(values = jenks_colors,
                    labels = formatted_labels) +         # Apply the viridis colors
  theme_tufte2() +
  theme(axis.text.x = element_text(color = "black", size = 28, angle = 0),
        axis.text.y = element_text(color = "black", size = 28, angle = 0),  
        axis.title.x = element_text(color = "black", size = 28, angle = 0),
        axis.title.y = element_text(color = "black", size = 28, angle = 90), 
        legend.title = element_text(size = 28),
        legend.text = element_text(size = 28),
        plot.title = element_text(size = 32)) +
  labs(title = "Distribution of census population in spatial units (LADs)",
       fill = "Jenks Breaks")

# ONLY UNCOMMENT IF WANT HISTOGRAM FOR WHOLE POPULATION. WILL ALSO NEED TO MODIFY CODE ABOVE TO MAKE HISTOGRAM OF df_bias$pop INSTEAD OF df_bias$pop_dfd
# ggsave(paste0(wd_local, "/bias-detection/outputs/preliminary/all/size-bias/histograms/population_lad.png"), bg="white")


if (dfd == "twitter") {
  ggsave(paste0(wd_local, "/bias-detection/outputs/preliminary/twitter/size-bias/histograms/twitter_lad_population.png"), bg="white")
  } else if (dfd == "fb_tts_census_week") {
  ggsave(paste0(wd_local, "/bias-detection/outputs/preliminary/fb/size-bias/histograms/fb_tts_lad_census_week_population.png"), bg="white")
  } else if (dfd == "fb_tts_census_month") {
  ggsave(paste0(wd_local, "/bias-detection/outputs/preliminary/fb/size-bias/histograms/fb_tts_lad_census_month_population.png"), bg="white")
  } else if (dfd == "fb_stt_census_week") {
  ggsave(paste0(wd_local, "/bias-detection/outputs/preliminary/fb/size-bias/histograms/fb_stt_lad_census_week_population.png"), bg="white")
  } else if (dfd == "fb_stt_census_month") {
  ggsave(paste0(wd_local, "/bias-detection/outputs/preliminary/fb/size-bias/histograms/fb_stt_lad_census_month_population.png"), bg="white")
  }
```

## Histogram bias

```{r}
# Compute the breaks using Jenks natural breaks
num_classes <- 8  # Specify how many breaks you want

df_bias$dropna <- df_bias$bias

# UNCOMMENT IF COMPUTING HISTOGRAM OF BIAS FOR TWITTER AND WANT TO REMOVE CITY OF LONDON LAD
# df_bias[df_bias$code == "E09000001", "dropna"] <- NA

# Create a data frame with histogram information
breaks <- classIntervals(df_bias$dropna, n = num_classes, style = "jenks")$brks
hist_data <- data.frame(value = df_bias$dropna)

# Create a factor variable for color breaks
hist_data$bins <- cut(hist_data$value, breaks = breaks, include.lowest = TRUE)

# Create a custom color palette from viridis
colors <- viridis(length(breaks) - 1)

# Manually format the legend labels to avoid scientific notation
formatted_labels <- sapply(1:(length(breaks) - 1), function(i) {
  paste0("[", format(breaks[i], scientific = FALSE), ", ", 
         format(breaks[i + 1], scientific = FALSE), ")")
})

# Plot the histogram with ggplot
ggplot(hist_data, aes(x = value, fill = bins)) +
  geom_histogram(color = "black", size = 0.3, bins=60) +  # Adjust binwidth as needed
  scale_fill_manual(values = colors,
                    labels = formatted_labels) +         # Apply the viridis colors
  theme_tufte2() +
  theme(axis.text.x = element_text(color = "black", size = 28, angle = 0),
        axis.text.y = element_text(color = "black", size = 28, angle = 0),  
        axis.title.x = element_text(color = "black", size = 28, angle = 0),
        axis.title.y = element_text(color = "black", size = 28, angle = 90), 
        legend.title = element_text(size = 28),
        legend.text = element_text(size = 28),
        plot.title = element_text(size = 32)) +
  labs(title = "Distribution of census population in spatial units (LADs)",
       fill = "Jenks Breaks")

# ONLY UNCOMMENT IF WANT HISTOGRAM FOR WHOLE POPULATION. WILL ALSO NEED TO MODIFY CODE ABOVE TO MAKE HISTOGRAM OF df_bias$pop INSTEAD OF df_bias$pop_dfd
# ggsave(paste0(wd_local, "/bias-detection/outputs/preliminary/all/size-bias/histograms/population_lad.png"), bg="white")


if (dfd == "twitter") {
  ggsave(paste0(wd_local, "/bias-detection/outputs/preliminary/twitter/size-bias/histograms/twitter_lad_bias_drop_E09000001.png"), bg="white")
  } else if (dfd == "fb_tts_census_week") {
  ggsave(paste0(wd_local, "/bias-detection/outputs/preliminary/fb/size-bias/histograms/fb_tts_lad_census_week_bias.png"), bg="white")
  } else if (dfd == "fb_tts_census_month") {
  ggsave(paste0(wd_local, "/bias-detection/outputs/preliminary/fb/size-bias/histograms/fb_tts_lad_census_month_bias.png"), bg="white")
  } else if (dfd == "fb_stt_census_week") {
  ggsave(paste0(wd_local, "/bias-detection/outputs/preliminary/fb/size-bias/histograms/fb_stt_lad_census_week_bias.png"), bg="white")
  } else if (dfd == "fb_stt_census_month") {
  ggsave(paste0(wd_local, "/bias-detection/outputs/preliminary/fb/size-bias/histograms/fb_stt_lad_census_month_bias.png"), bg="white")
  }
```


# Map bias

## England and Wales

### Static map

```{r}

st_bias_map <- df_boundaries %>% left_join(df_bias, by = "code") %>% st_as_sf() %>% filter(!grepl("^S", code))

st_bias_map[st_bias_map$code == "E09000001", "map_bias"] <- NA

breaks <- classIntervals(st_bias_map$map_bias, n = 8, style = "jenks")

st_bias_map$jenks <- cut(st_bias_map$map_bias, breaks$brks, include.lowest = TRUE)

ggplot(data = st_bias_map) +
  geom_sf(aes(fill = jenks)) +
  scale_fill_viridis_d(name = "Size of bias") +  # Adjust color palette
  labs(title = " ") +
  theme_map() +
  theme(
    legend.text = element_text(size = 24),       # Increase legend text size
    legend.title = element_text(size = 30),      # Increase legend title size
    legend.position = "right"                    # Optional: adjust legend position
  )


if (dfd == "twitter") {
  ggsave(paste0(wd_local, "/bias-detection/outputs/preliminary/twitter/size-bias/twitter_lad.png"), bg="white")
  } else if (dfd == "fb_tts_census_week") {
  ggsave(paste0(wd_local, "/bias-detection/outputs/preliminary/fb/size-bias/fb_tts_lad_census_week.png"), bg="white")
  } else if (dfd == "fb_tts_census_month") {
  ggsave(paste0(wd_local, "/bias-detection/outputs/preliminary/fb/size-bias/fb_tts_lad_census_month.png"), bg="white")
  } else if (dfd == "fb_stt_census_week") {
  ggsave(paste0(wd_local, "/bias-detection/outputs/preliminary/fb/size-bias/fb_stt_lad_census_week.png"), bg="white")
  } else if (dfd == "fb_stt_census_month") {
  ggsave(paste0(wd_local, "/bias-detection/outputs/preliminary/fb/size-bias/fb_stt_lad_census_month.png"), bg="white")
  }
```

### Interactive map

```{r}
tmap_mode("view") # enable interactivity
interactive_map <- tm_shape(st_bias_map) + # input data
    tm_borders() +
    tm_fill("map_bias", # draw and fill polygons
          palette = "viridis",
          title = "Size of bias",
          id="name")

if (dfd == "twitter"){
  tmap_save(interactive_map, paste0(wd_local, "/bias-detection/outputs/preliminary/twitter/size-bias/interactive/twitter_lad.html"))
  } else if (dfd == "fb_tts_census_week") {
  tmap_save(interactive, paste0(wd_local, "/bias-detection/outputs/preliminary/fb/size-bias/interactive/fb_tts_lad_census_week.html"))
  } else if (dfd == "fb_tts_census_month") {
  tmap_save(interactive, paste0(wd_local, "/bias-detection/outputs/preliminary/fb/size-bias/interactive/fb_tts_lad_census_month.html"))
  } else if (dfd == "fb_stt_census_week") {
  tmap_save(interactive, paste0(wd_local, "/bias-detection/outputs/preliminary/fb/size-bias/interactive/fb_stt_lad_census_week.html"))
  } else if (dfd == "fb_stt_census_month") {
  tmap_save(interactive, paste0(wd_local, "/bias-detection/outputs/preliminary/fb/size-bias/interactive/fb_stt_lad_census_month.html"))
  }

interactive_map
```

## London

### Static map

```{r}

st_bias_map <- df_boundaries %>% left_join(df_bias, by = "code") %>% st_as_sf() %>% filter(grepl("^E09", code)) 

st_bias_map[st_bias_map$code == "E09000001", "map_bias"] <- NA

st_bias_map$jenks <- cut(st_bias_map$map_bias, breaks$brks, include.lowest = TRUE)

ggplot(data = st_bias_map) +
  geom_sf(aes(fill = jenks)) +
  scale_fill_viridis_d(name = "Size of bias") +  # Adjust color palette
  labs(title = " ") +
  theme_map() +
  theme(
    legend.text = element_text(size = 24),       # Increase legend text size
    legend.title = element_text(size = 30),      # Increase legend title size
    legend.position = "right"                    # Optional: adjust legend position
  )

if (dfd == "twitter"){
  ggsave(paste0(wd_local, "/bias-detection/outputs/preliminary/twitter/size-bias/twitter_lad_london.png"), bg="white")
  } else if (dfd == "fb_tts_census_week") {
  ggsave(paste0(wd_local, "/bias-detection/outputs/preliminary/fb/size-bias/fb_tts_lad_london_census_week.png"), bg="white")
  } else if (dfd == "fb_tts_census_month") {
  ggsave(paste0(wd_local, "/bias-detection/outputs/preliminary/fb/size-bias/fb_tts_lad_london_census_month.png"), bg="white")
  } else if (dfd == "fb_stt_census_week") {
  ggsave(paste0(wd_local, "/bias-detection/outputs/preliminary/fb/size-bias/fb_stt_lad_london_census_week.png"), bg="white")
  } else if (dfd == "fb_stt_census_month") {
  ggsave(paste0(wd_local, "/bias-detection/outputs/preliminary/fb/size-bias/fb_stt_lad_london_census_month.png"), bg="white")
  }
```

### Interactive map

```{r}
tmap_mode("view") # enable interactivity
interactive_map <- tm_shape(st_bias_map) + # input data
    tm_borders() +
    tm_fill("map_bias", # draw and fill polygons
          palette = "viridis",
          title = "Size of bias",
          id="name")

if (dfd == "twitter"){
  tmap_save(interactive_map, paste0(wd_local, "/bias-detection/outputs/preliminary/twitter/size-bias/interactive/twitter_lad_london.html"))
  } else if (dfd == "fb_tts_census_week") {
  tmap_save(interactive_map, paste0(wd_local, "/bias-detection/outputs/preliminary/fb/size-bias/interactive/fb_tts_lad_london_census_week.html"))
  } else if (dfd == "fb_tts_census_month") {
  tmap_save(interactive_map, paste0(wd_local, "/bias-detection/outputs/preliminary/fb/size-bias/interactive/fb_tts_lad_london_census_month.html"))
  } else if (dfd == "fb_stt_census_week") {
  tmap_save(interactive_map, paste0(wd_local, "/bias-detection/outputs/preliminary/fb/size-bias/interactive/fb_stt_lad_london_census_week.html"))
  } else if (dfd == "fb_stt_census_month") {
  tmap_save(interactive_map, paste0(wd_local, "/bias-detection/outputs/preliminary/fb/size-bias/interactive/fb_stt_lad_london_census_month.html"))
  }

interactive_map
```
