---
title: 
"Final File of the Explain Bias Process. 

This file works for the following resources: FB TTS, FB STT, the three telephone data sets, and any other that complies with the input file format. 

Input files must be provided in a folder named Inputs::Population data(csv), Geographical areas (Municipalities & States (shp)), Bias data (csv),Labels (csv), Optimal parameters (optional) (cvs) 

The output files are saved in the folder specified in the code as Outputs (this can be renamed by changing the value of the variable).

Data                : Facebook (TTS y STT) and the three telephone data sets. 
Validation Method   : Blocked CV (5)
Model               : XGBOOST
graphics            : Included
Gamma               : No
Grid Search         : 5184
Last update         : 101225

"
format: html
editor: visual
---

## 1.Define / Install / Load libraries

Clear work space

```{r}
rm(list=ls())
```

```{r}
# Define packages
packages <- c(
  "xgboost",
  "parallel",
  "recommenderlab",
  "doParallel",
  "rsample",
  "tidyverse",
  "sf",
  "moments",
  "SHAPforxgboost",
  "shapviz",
  "pdp",
  "rsample",
  "foreach",
  "patchwork",
  "viridis",
  "scales",
  "tidyplots",
  "Ckmeans.1d.dp",
  "vtreat",
  "dplyr",
  "ggplot2",
  "readr",
  "ggthemes",
  "showtext",
  "Metrics",
  "caret",
  "yardstick"
)
# Install packages not yet installed
installed_packages <- packages %in% rownames(installed.packages())
if (any(installed_packages == FALSE)) {
  install.packages(packages[!installed_packages])
} 
# Load packages
invisible(lapply(packages, library, character.only = TRUE))
```

Dependences

```{r}
#| warning: false
#| message: false
# library necessary packages
library(xgboost)
library(parallel)
library(recommenderlab)
library(doParallel)
library(rsample)
library(tidyverse)
library(sf)
library(moments)
library(SHAPforxgboost)
library(shapviz)
library(pdp)
library(rsample)
library(foreach)
library(patchwork)
library(viridis)
library(scales)
library(tidyplots)
library(Ckmeans.1d.dp)
library(vtreat)
library(dplyr)
library(readr)
library(ggthemes)
library(showtext)
library(Metrics)
library(caret)
library(yardstick)
```

Set theme

```{r}
#| include = FALSE
source("../style/data-visualisation_theme.R")
```

## 2.Parameters

```{r}
# Folder where the outputs will be saved
Outputs <- "Final_Results_BlockedCV_GS5184_P05_Knn"
# Value the training set from 0 to 1
# train_proportion <- 0.7
VFolds <- 5
# Paths
Path01 <- "Inputs/Census_data.csv"
Path02 <- "Inputs/SHP/MunicipalitiesMX.shp"
Path03 <- "Inputs/Active_population_bias_knn_p0005.csv"
Path04 <- "Inputs/Labels.csv" # TTS
Path05 <- "Inputs/SHP/StatesMX.shp"
# Path06 <- "Inputs/optimal_parameters_knn.csv"

print(Path01)
print(Path02)
print(Path03)
print(Path04)
print(Path05)
# print(Path06)
```

## 3.Data

### 3.1 Population data

```{r}
Population_data <- read.csv(Path01,fileEncoding = "latin1")
colnames(Population_data)[1] <- "Area_code"
colnames(Population_data)[2] <- "Area_name"
colnames(Population_data)[3] <- "Total_Population"

# Verify Population data
head(Population_data,n=5)
# Verify missing values
print(t(t(colSums(is.na(Population_data)))))
# Verify dimensions
print(dim(Population_data))

# Population labels  ==============================================================
Population_data_labels <- names(Population_data)
```

### 3.2 Geographical areas (Municipalities) & Boundaries

```{r}
# Geographical areas ==============================================================
Geographical_areas <- st_read(Path02)
colnames(Geographical_areas)[1] <- "Area_code"
Geographical_areas <-  Geographical_areas %>% select(Area_code, geometry)

# Verify data
head(Geographical_areas,n=5)
# Verify missing values
print(t(t(colSums(is.na(Geographical_areas)))))
# Verify dimensions
print(dim(Geographical_areas))

# Boundaries ======================================================================
Boundaries <- Geographical_areas
# Simplifies the complexity of spatial geometries
Boundaries <- Boundaries %>%st_simplify(preserveTopology = TRUE, dTolerance = 1000)
# Transforms the coordinate system of spatial geometries
Boundaries <- st_transform(Boundaries,  crs = 4326)

# Verify data
head(Boundaries,n=5)
# Verify missing values
print(t(t(colSums(is.na(Boundaries)))))
# Verify dimensions
print(dim(Boundaries))
```

### 3.3 Bias data

```{r}
Bias_data <- read.csv(Path03,fileEncoding = "latin1")
colnames(Bias_data)[1] <- "Area_code"
print(head(Bias_data,n=5))
```

### 3.4 Labels

```{r}
Labels <- read.csv(Path04,fileEncoding = "latin1", header = TRUE)
colnames(Labels)[1] <- "Code_label"
colnames(Labels)[2] <- "Plot_label"
print(Labels)
```

### 3.5 Geographical areas (States)

```{r}
Geographical_areas_states <- st_read(Path05)
Geographical_areas_states$Area_code <- 1:32
colnames(Geographical_areas_states)[4] <- "Area"
Geographical_areas_states <-  Geographical_areas_states %>% 
  select(Area_code, geometry)

# Verify data
head(Geographical_areas_states,n=5)
# Verify missing values
print(t(t(colSums(is.na(Geographical_areas_states)))))
# Verify dimensions
print(dim(Geographical_areas_states))
```

### 3.6 Optimal parameters (if exists)

```{r}
# We check if the variable exists
if (exists("Path06")){
    # We check if the PATH is valid.We use file.exists() if it is a CSV file
    if (file.exists(Path06)){ 
        optimal_pars <- read.csv(Path06, fileEncoding = "latin1")
        print("Parameters loaded successfully:")
        print(optimal_pars)
    }else{
        # The variable exists, but the PATH is incorrect or the file does not exist
        print(paste("WARNING: The path",Path06," does not contain a valid file.There are no previous parameters."))
    }
}else{
    # The variable 'Path06' does NOT exist in the environment.
    print("WARNING: The variable Path06 does not exist in the environment. There are no previous parameters.")
}
```

## 4.Data wrangling

### Extract centroids

```{r}
# Calculate the centroid of each geometry within the Geographical_areas object.
centroids_df <- st_centroid(Geographical_areas) %>% 
  st_coordinates() %>% 
  as.data.frame() %>% 
  rowid_to_column("row_id")

# Combine Geographical_areas with centroids_df
centroids_with_ids <- Geographical_areas %>%
  st_set_geometry(NULL) %>% # Remove geometry to avoid issues with join
  rowid_to_column("row_id") %>% # Create a temporary row_id for merging
  cbind(centroids_df) 

# Remove row_id and rename x,y
centroids_with_ids <- centroids_with_ids %>% 
  select(-c(row_id)) %>% 
  rename(
    longitude = "X",
    latitude = "Y"
  )

# Casting "Area_code" as integer in order to use left_join
centroids_with_ids <- centroids_with_ids %>%
  mutate(Area_code = as.integer(Area_code))

# Create df_extended by join Population_data and centroids_with_ids
df_extended <- full_join(Population_data, centroids_with_ids, by = join_by(Area_code == Area_code) ) 

print(names(df_extended))
```

### Add bias_w (Add spatial weights If they are included as a feature)

```{r}
# Add spatial weights to the label data frame and df_extended if they are included 
if ("bias_w" %in% names(Bias_data)) {
  Population_data_labels <- c(Population_data_labels, "bias_w")
  df_extended <- merge(df_extended, Bias_data[, c("Area_code", "bias_w")], by = "Area_code")
}
print(Population_data_labels)
# Verify df_extended
head(df_extended,n=5)
# Verify missing values
print(t(t(colSums(is.na(df_extended)))))
# Verify dimensions
print(dim(df_extended))
```

### 4.1 Standarization

```{r}
# select numeric columns and standardize them
df_standardised <- df_extended %>%
  dplyr::select(-c(Area_code)) %>% 
  mutate(across(where(is.numeric), ~ (. - mean(.)) / sd(.))) %>%
  cbind(df_extended %>% dplyr::select(Area_code))

# Verify df_standardised
df_standardised
print(t(t(colSums(is.na(df_standardised)))))
print(dim(df_standardised))
```

### 4.2 Add bias (Add bias as dependent variable)

```{r}
input_df <- merge(df_standardised, Bias_data[, c("Area_code", "bias")], by = "Area_code")

# Verify input_df
head(input_df,n=5)
print(t(t(colSums(is.na(input_df)))))
print(dim(input_df))
```

### 4.3 Missing values

```{r}
# Check for missing values
print(t(t(colSums(is.na(input_df)))))

print(which(is.na(input_df), arr.ind = TRUE))
input_df <- input_df %>% 
  drop_na()

print(dim(input_df))
```

Removing unnecessary objects (optional)

```{r}
# rm(Geographical_areas, df_standardised,df_extended)
print("Removing unnecessary objects (optional)")
```

## 5.Setting up model

### Create folders for outputs and saving

```{r}
# 1. Create the main folder if it does not exist
if (!dir.exists(Outputs)) {
  dir.create(Outputs)
  cat(paste0("Main Folder Output '", Outputs, "' successfully created.\n"))
} else {
  cat(paste0("Main Folder Output '", Outputs, "' already exists.\n"))
}

# 2. Create the subfolders inside the main folder.
Subfolders <- c("Inputs","01.Feature Importance","02.Shap Plot","03.Shap Dependence","05.Map","Predictions")
for (Subfolder in Subfolders) {
  path_subfolder <- file.path(Outputs, Subfolder)
  if (!dir.exists(path_subfolder)) {
    dir.create(path_subfolder)
    cat(paste0("Subfolder '", Subfolder, "' created in '", Outputs, "' successfully.\n"))
  } else {
    cat(paste0("Subfolder '", Subfolder, "' already exists in '", Outputs, "'.\n"))
  }
}
cat("Folder creation process completed.\n")

# SAVING ==========================================================================
# =================================================================================
# INPUT_DF ========================================================================
# Try to write the CSV file and save the result.
write_output <- try(write.csv(input_df, file.path(Outputs, "Inputs/Input_df.csv"), fileEncoding = "latin1", row.names = FALSE))
# Check if there were any errors during writing
if (inherits(write_output, "try-error")) {
  cat(paste0("Error saving: 'Input_df.csv' in '", Outputs, "'.\n"))
  cat("Error message: ", as.character(write_output), "\n")
} else {
  cat(paste0("Input_df.csv successfully created in '", Outputs, "/Inputs\n"))
}

# POPULATION DATA =================================================================
# Try to write the CSV file and save the result.
write_output <- try(write.csv(Population_data, file.path(Outputs, "Inputs/Population_data.csv"), fileEncoding = "latin1", row.names = FALSE))
# Check if there were any errors during writing
if (inherits(write_output, "try-error")) {
  cat(paste0("Error saving: 'Population_data.csv' in '", Outputs, "/Inputs'.\n"))
  cat("Error message: ", as.character(write_output), "\n")
} else {
  cat(paste0("Population_data.csv successfully created in '", Outputs, "/Inputs\n"))
}

# GEOGRAPHICAL AREAS ==============================================================
# Try to write the SHP file and save the result.
write_output <- try(st_write(Geographical_areas, file.path(Outputs, "Inputs/Geographical_areas.shp"), row.names = FALSE))
# Check if there were any errors during writing
if (inherits(write_output, "try-error")) {
  cat(paste0("Error saving: 'Geographical_areas.shp' in '", Outputs, "'.\n"))
  cat("Error message: ", as.character(write_output), "\n")
} else {
  cat(paste0("Geographical_areas.shp successfully created in '", Outputs, "/Inputs\n"))
}

# BIAS DATA =======================================================================
# Try to write the CSV file and save the result.
write_output <- try(st_write(Bias_data, file.path(Outputs, "Inputs/Bias_data.csv"), fileEncoding = "latin1", row.names = FALSE))
# Check if there were any errors during writing
if (inherits(write_output, "try-error")) {
  cat(paste0("Error saving: 'Bias_data.csv' in '", Outputs, "'.\n"))
  cat("Error message: ", as.character(write_output), "\n")
} else {
  cat(paste0("Bias_data.csv successfully created in '", Outputs, "/Inputs\n"))
}

# LABELS ==========================================================================
# Try to write the CSV file and save the result.
write_output <- try(st_write(Labels, file.path(Outputs, "Inputs/Labels.csv"), fileEncoding = "latin1", row.names = FALSE))
# Check if there were any errors during writing
if (inherits(write_output, "try-error")) {
  cat(paste0("Error saving: 'Labels.csv' in '", Outputs, "'.\n"))
  cat("Error message: ", as.character(write_output), "\n")
} else {
  cat(paste0("Labels.csv successfully created in '", Outputs, "/Inputs\n"))
}

# GEOGRAPHICAL AREAS STATES =======================================================
# Try to write the SHP file and save the result.
write_output <- try(st_write(Geographical_areas_states, file.path(Outputs, "Inputs/Geographical_areas_states.shp"), row.names = FALSE))
# Check if there were any errors during writing
if (inherits(write_output, "try-error")) {
  cat(paste0("Error saving: 'Geographical_areas_states.shp' in '", Outputs, "'.\n"))
  cat("Error message: ", as.character(write_output), "\n")
} else {
  cat(paste0("Geographical_areas_states.shp successfully created in '", Outputs, "/Inputs\n"))
}
```

### 5.1 Dependent and independent variables

```{r}
# cut <- "random"

variables_to_eliminate <- c(
  "Area_code",
  "Area_name",
  "longitude", 
  "latitude",
  "bias")

print("=======================================================================")
print("DISCARDED VARIABLES:")
print("-----------------------------------------------------------------------")
print(variables_to_eliminate)

xvars <- colnames(input_df)

x <- xvars[!(xvars %in% variables_to_eliminate)]
y <- "bias"

# Independent variables
print("=======================================================================")
print("INDEPENDENT VARIABLES:")
print("-----------------------------------------------------------------------")
print(x)

# Dependent variable
print("=======================================================================")
print("DEPENDENT VARIABLES:")
print("-----------------------------------------------------------------------")
print(y)
```

### 5.2 Validation Method - Defining training and test data sets

### **Blocked CV**

```{r}
# install.packages("spatialsample")
library(rsample)
library(sf)
library(spatialsample)

# Convert to spatial format
input_sf <- st_as_sf(input_df, coords = c("longitude", "latitude"), crs = 4326)

# Generating seed
set.seed(123)

# Data Division
blocked_folds <- spatial_block_cv(
  input_sf,
  v = VFolds
)

# Partition structure
print(blocked_folds)

for (k in 1:VFolds)
{
print(paste("---------------- Fold:", k, " begins----------------"))
  
# Get the first fold
fold <- blocked_folds$splits[[k]]

# Get the training Set
train_sample <- training(fold)

# Get the test Set
test_sample <- testing(fold)

# Verify train_sample
#head(train_sample,n=5)
print(dim(train_sample))

# Verify test_sample
#head(test_sample,n=5)
print(dim(test_sample))
}
```

### A) Treatment plan - Prepare data

```{r}
library(sf)
library(vtreat)
library(rsample)

# List to store the vtreat transformed datasets for all folds
vtreat_processed_folds <- list()

# Iterate over each fold
for (k in 1:VFolds){
# for (k in 1:5){
  
  print(paste("--- Processing Fold:", k, "---"))
  
  # Get the current split
  current_split <- blocked_folds$splits[[k]]
  
  # Get Training and Test indices
  ##train_indices <- current_split$in_id
  ##all_indices   <- 1:nrow(input_df)
  ##test_indices  <- all_indices[!(all_indices %in% train_indices)]
  
  # Separate the original dataframe (input_df) using the indices
  ##train_sample <- input_df[train_indices, ]
  ##test_sample  <- input_df[test_indices, ]
  # print(head(test_sample))
  
  # Get Training and Test
  train_sample <- rsample::training(current_split)
  test_sample  <- rsample::testing(current_split)

  # Create the treatment plan from the training data
  treatplan <- vtreat::designTreatmentsZ(
    dframe       = train_sample, 
    varlist      = x,       # Predictor (independent) variables
    verbose = FALSE
  )
  
  # Get the "clean" variable names from the scoreFrame
  new_vars <- treatplan %>%
  magrittr::use_series(scoreFrame) %>%        
  dplyr::filter(code %in% c("clean", "lev")) %>% 
  magrittr::use_series(varName)  

  # Prepare the training data
  features_train <- vtreat::prepare(
    treatplan,           
    train_sample,        
    varRestriction = new_vars
  ) %>% as.matrix()
  
  response_train <- train_sample[,y]

  # Prepare the test data
  features_test <- vtreat::prepare(
    treatplan,           
    test_sample,        
    varRestriction = new_vars
  ) %>% as.matrix()

  response_test <- test_sample[,y]
  
  # Store the processed results for this fold
  vtreat_processed_folds[[k]] <- list(
    features_train = features_train,
    response_train = response_train,
    features_test  = features_test,
    response_test  = response_test
  )

  print(paste("Treatment plan applied to Fold", k, 
              "- Dimension of features_train:", dim(features_train)[1], "x", dim(features_train)[2]))
}

# --- VERIFICATION OF RESULTS ---
# vtreat_processed_folds now contains 5 elements ready for modeling
primer_fold <- vtreat_processed_folds[[1]]
print("Processed variables from the first fold ready for modeling:")

# Independent Variables
print("INDEPENDENT VARIABLES:")
print("=======================================================================")
print(colnames(primer_fold$features_train))
print("=======================================================================")
print(length(primer_fold$response_train))
```

### 5.3 Tuning - Hypergrid Parameters (Only if the optimal parameters are not available)

#### 2) Parameters

```{r}
# Create hyperparameter grid

# ORIGINAL PARAMETERS
 hyper_grid_O <- expand.grid(
  #nrounds         = c(50, 100, 150),     # Number of boosting rounds
  eta              = c(.01, .05, .1, .3), # Learning rate
  #gamma           = c(0, 1, 5),          # Minimum loss reduction
  max_depth        = c(1, 3, 5, 7),       # Maximum depth of a tree
  min_child_weight = c(1, 3, 5, 7),       # Minimum sum of instance weight
  subsample        = c(.65, .8, 1),       # Fraction of rows per tree
  colsample_bytree = c(.8, .9, 1),        # Fraction of columns per tree
  lambda           = c(0, 0.5, 1),        # L2 regularization term (Ridge)
  alpha            = c(0, 0.5, 1),        # L1 regularization term (LASSO)
  optimal_trees    = 0,                   # a place to dump results
  min_RMSE         = 0                    # a place to dump results
)


# TEST PARAMETERS
 hyper_grid_T <- expand.grid(
 #nrounds          = c(50, 100, 150),     # Number of boosting rounds
 eta               = c(.01, .05),         # Learning rate
 #gamma            = c(0, 1, 5),          # Minimum loss reduction
 max_depth         = c(1, 3),             # Maximum depth of a tree
 min_child_weight  = c(1),                # Minimum sum of instance weight
 subsample         = c(.65, .8),          # Fraction of rows per tree
 colsample_bytree  = c(.8),               # Fraction of columns per tree
 lambda            = c(0, 0.5),           # L2 regularization term (Ridge)
 alpha             = c(0),           # L1 regularization term (LASSO)
 #alpha             = c(0, 0.5),           # L1 regularization term (LASSO)
 optimal_trees     = 0,                   # a place to dump results
 min_RMSE          = 0                    # a place to dump results
)

# SELECT PARAMETERS
hyper_grid <- hyper_grid_O # ORIGINAL
# hyper_grid <- hyper_grid_T # TEST

nrow(hyper_grid)

# HYPERGRID ==========================================================================
# Try to write the CSV file and save the result.
write_output <- try(write.csv(hyper_grid, file.path(Outputs, "/Hypergrid.csv"), fileEncoding = "latin1", row.names = FALSE))
# Check if there were any errors during writing
if (inherits(write_output, "try-error")) {
  cat(paste0("Error saving: 'Hypergrid.csv' in '", Outputs, "'.\n"))
  cat("Error message: ", as.character(write_output), "\n")
} else {
  cat(paste0("Hypergrid.csv successfully created in '", Outputs, "\n"))
}
```

### 5.4 Parallelised code

We apply a loop procedure to loop through and apply a XGBoost model for each hyperparameter combination and dump the results in the hyper_grid data frame. This is done in a paralellised way.

#### **Blocked CV**

Selecting optimal parameters

```{r}
# List for optimal parameters
optimal_pars_fold <- list()

for (k in 1:VFolds){
  
  # Define file name and path
  filename <- paste0("Results_hypergrid_fold_", k, ".csv")
  file_path <- file.path(Outputs, filename)
  
  # Read file
  fold_results <- tryCatch({
        read_csv(file_path, show_col_types = FALSE)
    }, error = function(e) {
        cat(paste0("Error reading file for Fold ", k, ": ", e$message, "\n"))
        return(NULL) # Returns NULL if there is an error
    })
  
  # Verify and extract the best set of parameters
    if (!is.null(fold_results)) {
        # Find the row with the minimum RMSE for this fold
        optimal_pars_k <- fold_results %>% 
            dplyr::slice_min(order_by = min_RMSE, n = 1) %>%
            # Ensure the fold_id column is correct if needed
            dplyr::mutate(fold_id = k)
        
        # Store the result in the list
        optimal_pars_fold[[k]] <- optimal_pars_k
        
        cat(paste0("Fold ", k, " processed. Min RMSE: ", round(optimal_pars_k$min_RMSE, 5), "\n"))
    }
  
}

# Convert the list of 5 dataframes (1 row each) into a single dataframe of 5 rows
optimal_pars_folds <- dplyr::bind_rows(optimal_pars_fold)

# The dataframe 'final_optimal_params_df' now contains the 5 sets of optimal parameters (one per fold).
print(paste0("Parámetros Óptimos consolidados para los ",k," Folds:"))

# RESULTS =========================================================================
# Try to write the CSV file and save the result.
write_output <- try(write.csv(optimal_pars_folds, file.path(Outputs, "/Optimal_parameters.csv"), fileEncoding = "latin1", row.names = FALSE))
# Check if there were any errors during writing
if (inherits(write_output, "try-error")) {
  cat(paste0("Error saving: 'Optimal_parameters.csv' in '", Outputs, "'.\n"))
  cat("Error message: ", as.character(write_output), "\n")
} else {
  cat(paste0("Optimal_parameters.csv successfully created in '", Outputs, "\n"))
}

print(optimal_pars_folds)
```

## 6.Final model

Once you've found the optimal model, we can fit our final model

### 6.1 Fit final model, Predictions XGBoost & Metric tables

```{r}
# Initialize lists to store final models and metrics
final_models <- list()
all_fold_metrics <- list()

# -----------------------------------------------------------
## FINAL LOOP: Train and Evaluate the K Models
# -----------------------------------------------------------
for (k in 1:VFolds) {
    
    # -----------------------------------------------------------
    ## 6.1 FIT FINAL MODEL
    # -----------------------------------------------------------
    print(paste("--- Training Final Model for Fold:", k, " ---"))
  
    # 1. Get the data and parameters for the current Fold
    current_data   <- vtreat_processed_folds[[k]]
    optimal_pars_k <- optimal_pars_folds[k, ] 
    
    # 2. Define Data and Parameters
    X_train_mat <- current_data$features_train
    y_train_vec <- as.numeric(current_data$response_train[[1]]) 
    dtrain <- xgboost::xgb.DMatrix(data = X_train_mat, label = y_train_vec)
    
    # Parameters (Extracted from the optimal dataframe)
    params <- list(
        eta              = optimal_pars_k$eta,
        max_depth        = optimal_pars_k$max_depth,
        min_child_weight = optimal_pars_k$min_child_weight,
        subsample        = optimal_pars_k$subsample,
        colsample_bytree = optimal_pars_k$colsample_bytree,
        lambda           = optimal_pars_k$lambda,
        alpha            = optimal_pars_k$alpha,
        objective        = "reg:squarederror"
    ) 
    optimal_nrounds <- optimal_pars_k$optimal_trees
    
    # 3. Train final model
    xgb.fit.final <- xgboost::xgb.train(
        params  = params,
        data    = dtrain,
        nrounds = optimal_nrounds,
        verbose = 0
        #seed    = 123
    )
    
    print(paste("Fold", k, "Model trained with", optimal_nrounds, "trees."))
    
    # 4. Store and Save the Model
    # Store the model in the model list
    final_models[[k]] <- xgb.fit.final
    # Save model
    model_filename <- paste0("xgb-fit-final-model-fold-",k,".rds")
    saveRDS(xgb.fit.final, file.path(Outputs, model_filename))
    
    # -----------------------------------------------------------
    ## 6.2 TRAINING AND TEST PREDICTIONS
    # -----------------------------------------------------------
    print(paste("--- Training and Test Predictions for Fold:", k, " ---"))
    
    # 1. Define Test Data 
    X_test_mat <- current_data$features_test
    y_test_vec <- as.numeric(current_data$response_test[[1]]) 
    dtest <- xgboost::xgb.DMatrix(data = X_test_mat)
    
    # 2. Predictions
    # Train Predictions
    predictions_train_k <- predict(xgb.fit.final, dtrain)
    # Test Predictions
    predictions_test_k <- predict(xgb.fit.final, dtest)
    
    # -----------------------------------------------------------
    ## 6.2.1 SAVING TRAINING AND TEST PREDICTIONS
    # -----------------------------------------------------------
    # 1. Preparation of TRAINING data with all attributes
    train_features_df <- as.data.frame(X_train_mat) 
    # Create the final TRAINING results table
    final_train_df <- train_features_df %>%
        dplyr::mutate(
            Fold_ID = k,
            Observed_Value = y_train_vec,
            Predicted_Value = predictions_train_k
        )%>%
        # Reorder columns to have Fold_ID, Observed, and Predicted at the top
        dplyr::select(Fold_ID,Observed_Value,Predicted_Value, dplyr::everything())

    # 2. Preparation of TEST data with all attributes
    test_features_df <- as.data.frame(X_test_mat)
    # Create the final TEST results table
    final_test_df <- test_features_df %>%
        dplyr::mutate(
            Fold_ID = k,
            Observed_Value = y_test_vec,
            Predicted_Value = predictions_test_k
        )%>%
        # Reorder columns
        dplyr::select(Fold_ID,Observed_Value,Predicted_Value, dplyr::everything())

    # 3. Save as CSV with validation
    # --- 3a. Save TRAINING ---
    filename_train <- paste0("Predictions/Fold_",k,"_full_predictions_train.csv")
    file_path_train <- file.path(Outputs,filename_train)
    write_output_train<-try(readr::write_csv(final_train_df,file=file_path_train))
    if (inherits(write_output_train, "try-error")){
        cat(paste0("Error saving full_predictions_train for Fold ", k, ".\n"))
    }else{
        cat(paste0(filename_train, " successfully created with attributes.\n"))
    }
  
    # --- 3b. Save TEST ---
    filename_test <- paste0("Predictions/Fold_", k, "_full_predictions_test.csv")
    file_path_test <- file.path(Outputs, filename_test)
    write_output_test<-try(readr::write_csv(final_test_df, file = file_path_test))
    if (inherits(write_output_test, "try-error")) {
        cat(paste0("Error saving full_predictions_test para Fold ", k, ".\n"))
    }else{
        cat(paste0(filename_test, " successfully created with attributes.\n"))
    }
    
    # -----------------------------------------------------------
    ## 6.3 METRICS
    # -----------------------------------------------------------    
    # 3. Metrics
    # Auxiliary function to calculate R2
    r2_calc <- function(actual, predicted) {
        SS_res <- sum((predicted - actual)^2)
        SS_tot <- sum((actual - mean(actual))^2)
        return(1 - (SS_res / SS_tot))}
    
    # Calculate the error vector (Residues)
    errors_train_k <- predictions_train_k - y_train_vec
    errors_test_k  <- predictions_test_k - y_test_vec

    # -----------------------------------------------------------
    # 3A. Training Metrics
    # -----------------------------------------------------------
    metrics_train_k <- list(
    RMSE_Train = sqrt(mean((predictions_train_k - y_train_vec)^2)),
    Pearson_Train = cor(predictions_train_k, y_train_vec,method="pearson"),
    Spearman_Train=cor(predictions_train_k,y_train_vec,method= "spearman"),
    SD_Error_Train = sd(errors_train_k),
    R2_Train = r2_calc(y_train_vec, predictions_train_k))
    # -----------------------------------------------------------
    # 3B. Test Metrics
    # -----------------------------------------------------------
    metrics_test_k <- list(
    RMSE_Test = sqrt(mean((predictions_test_k - y_test_vec)^2)),
    Pearson_Test = cor(predictions_test_k, y_test_vec, method = "pearson"),
    Spearman_Test=cor(predictions_test_k, y_test_vec, method = "spearman"),
    SD_Error_Test = sd(errors_test_k),
    R2_Test = r2_calc(y_test_vec, predictions_test_k))

    # 4. Storage
    #Combine the Train and Test metrics into a single list for the fold
    metrics_k <- c(
      list(fold_id = k),
      metrics_train_k,
      metrics_test_k)

    # Store the metrics for this Fold in the global list
    all_fold_metrics <- append(all_fold_metrics, list(metrics_k))

    # Report
    # print(paste("Fold", k, "Evaluation: 
    #            | RMSE Train:", 
    #            round(metrics_k$RMSE_Train, 4), 
    #            "| RMSE Test:", 
    #            round(metrics_k$RMSE_Test, 4)))
    

    # -----------------------------------------------------------
    ## 6.3.1 METRICS TABLE
    # ----------------------------------------------------------- 
    # 1. Create metrics vectors
    metrics_list <- c("RMSE","Pearson","Spearman","SD_Error","R2")

    # 2. Extract the values from Training and Test
    train_values <- c(
    metrics_train_k$RMSE_Train,
    metrics_train_k$Pearson_Train,
    metrics_train_k$Spearman_Train,
    metrics_train_k$SD_Error_Train,
    metrics_train_k$R2_Train)

    test_values <- c(
    metrics_test_k$RMSE_Test,
    metrics_test_k$Pearson_Test,
    metrics_test_k$Spearman_Test,
    metrics_test_k$SD_Error_Test,
    metrics_test_k$R2_Test)

    # 3. Summary table
    summary_table_k <- data.frame(
    Metric = metrics_list,
    Train = train_values,
    Test = test_values)

    print(paste("Fold performance table", k, ":"))
    print(summary_table_k)

    # 4. Saving summary table
    filename <- paste0("Model-performance_fold_", k, ".csv")
    file_path <- file.path(Outputs, filename)

    # Validation
    write_output <- try(
    readr::write_csv(summary_table_k, file = file_path))

    if (inherits(write_output, "try-error")) {
        cat(paste0("Error saving performance table for Fold ", k, ".\n"))
    } else {
        cat(paste0("Model-performance_fold_", k, ".csv successfully created.\n"))
    }
    
    # -----------------------------------------------------------
    ## 6.4 GLOBAL METRICS
    # -----------------------------------------------------------
    if (k == VFolds)
    {
      # 6.4.1 Convert the list of results from the 5 folds into a dataframe
      consolidated_metrics_df <- dplyr::bind_rows(all_fold_metrics)
      
      # 6.4.2 Calculate the final return (Average AVG and SD of the RMSEs of the 5 folds)
      final_performance_summary <- consolidated_metrics_df %>%
      dplyr::summarise(
        # TEST METRICS
        Avg_RMSE_Test     = mean(RMSE_Test),
        SD_RMSE_Test      = sd(RMSE_Test), 
        Avg_SD_Error_Test = mean(SD_Error_Test),
        Avg_R2_Test       = mean(R2_Test),
        Avg_Pearson_Test  = mean(Pearson_Test),
        Avg_Spearman_Test = mean(Spearman_Test),
        
        # TRAINING METRICS 
        Avg_RMSE_Train     = mean(RMSE_Train),
        SD_RMSE_Train      = sd(RMSE_Train),
        Avg_SD_Error_Train = mean(SD_Error_Train),
        Avg_R2_Train       = mean(R2_Train),
        Avg_Pearson_Train  = mean(Pearson_Train),
        Avg_Spearman_Train = mean(Spearman_Train)
    )
    
    # 6.4.3 Transpose to obtain the summary table format (Metric, Train, Test)
    ordered_metrics <- c(
      "RMSE", 
      "Pearson", 
      "Spearman",
      "SD_Error", 
      "R2", 
      "SD_RMSE")  

    final_summary_table <- final_performance_summary %>%
    tidyr::pivot_longer(
        cols = starts_with("Avg_"), 
        names_to = c("Type", "Metric_Name", "Set"), 
        names_pattern = "(Avg)_(.*)_(Test|Train)", 
        values_to = "Value"
    ) %>%
    tidyr::pivot_wider(
        names_from = Set,
        values_from = Value
    ) %>%
    dplyr::filter(Metric_Name %in% c(
      "RMSE", 
      "R2", 
      "Pearson", 
      "Spearman", 
      "SD_Error")) %>%
    dplyr::select(Metric = Metric_Name, Train, Test) %>%
    
    dplyr::bind_rows(
        data.frame(
            Metric = "SD_RMSE", 
            Train = final_performance_summary$SD_RMSE_Train, 
            Test = final_performance_summary$SD_RMSE_Test
        )
    ) %>% dplyr::arrange(match(Metric, ordered_metrics))
    
    # 6.4.4 Save the Final Summary Table with Validation
    final_report_filename <- "Final_Blocked_CV_Performance.csv"
    file_path_final <- file.path(Outputs, final_report_filename)
    write_output_final <- try(readr::write_csv(final_summary_table, file = file_path_final))
    if (inherits(write_output_final, "try-error")) {
        cat(paste0("Fatal error saving the final report.\n"))
    } else {
        cat(paste0("FINAL REPORT successfully saved: ", final_report_filename, "\n"))
        print(final_summary_table)
    }
    }
  
}
print("Training and predictions of the final K models completed")
```

## 7.Graphics

### 7.1 Feature_Importance_PDF

#### Fold by Fold

```{r}
library(shapviz)
library(dplyr)
library(ggplot2)
library(scales)
library(showtext)

# Font
font <- "sans"
# font <- "Fira Sans"
# font <- "Source Sans Pro"
tamanio <- 17

# ---------------------------------------------------------------------------------
# 0. PREPARATION
# ---------------------------------------------------------------------------------
for (k in 1:VFolds) {
  # 1. Load the final trained model for Fold k
  model_filename <- paste0("xgb-fit-final-model-fold-", k, ".rds")
  file_path <- file.path(Outputs, model_filename)
  
    # Handling the error if the model does not exist
  xgb.fit.final <- tryCatch({
      readRDS(file_path)
  }, error = function(e) {
      cat(paste0("Error: Could not load model for Fold ", k, ".\n"))
      return(NULL)
  })
  # Skip to the next fold if the model does not exist
  if (is.null(xgb.fit.final)) next 
  
  # 2. Get the training set from Fold k (features_train)
  current_data <- vtreat_processed_folds[[k]]
  features_train <- current_data$features_train
# ---------------------------------------------------------------------------------
# 1.UNIQUE SHAP CALCULATION
# ---------------------------------------------------------------------------------
shp <- shapviz(
    xgb.fit.final, 
    X_pred = features_train, 
    X = features_train      
)
## print(shp)
## print(names(shp$X))

# ---------------------------------------------------------------------------------
# 2.DATA EXTRACTION, LABELING AND PREPARATION
# ---------------------------------------------------------------------------------
importance_data <- shapviz::sv_importance(shp, kind = "bar", max_display = 20L) %>%
    .$data %>% 
    dplyr::left_join(Labels, by = c("feature" = "Code_label")) %>% 
    dplyr::mutate(Fold_ID = k) %>% 
    dplyr::select(feature = Plot_label, value, Fold_ID)

## print(importance_data)

importance_data$feature <- factor(importance_data$feature, levels = rev(importance_data$feature))

## print(importance_data)
# ---------------------------------------------------------------------------------
# 3.GRAPH GENERATION
# ---------------------------------------------------------------------------------
Plot <- ggplot(importance_data, aes(x = value, y = feature))+
        geom_segment(aes(x = 0, 
                         xend = value),
                     color = "#000000", 
                     linewidth = 0.5, 
                     linetype = "solid") +
        geom_point(aes(color = value > 0), 
                   size = 7, 
                   alpha = 1) +
        scale_color_manual(values = c("TRUE" = "#800080", "FALSE" = "#800080")) +
        labs(
          title = "A) Feature importance",
          x     = "Impact SHAP Value",
          y     = NULL)+
        theme_classic()+
        scale_x_continuous(
          limits = c(0, max(importance_data$value) * 1.05), 
          expand = c(0, 0))+
        theme(
          plot.title = element_text(size   = tamanio, 
                                face   = "bold", 
                                hjust  = 0, 
                                color  = "#000000", 
                                family = font),
          plot.title.position = "plot",
          axis.title.x = element_text(size = tamanio),
          axis.text.x=element_text(size = tamanio,color = "#000000",family = font),
          axis.text.y=element_text(size = tamanio,color = "#000000",family = font),
          legend.position = "none",
          plot.margin = margin(5, 20, 5, 5)
    )
print(Plot)

# ---------------------------------------------------------------------------------
# 4. SAVING
# ---------------------------------------------------------------------------------
# CSV
importance_data_for_csv <- importance_data %>%
      dplyr::select(
        Fold_ID,
        Feature_Label = feature,
        Impact_SHAP_Value = value)

csv_file_name<- paste0("01.Feature Importance/Feature_Importance_Fold_", k, ".csv")
csv_full_path <- file.path(Outputs, csv_file_name)
write.csv(importance_data_for_csv, csv_full_path, row.names = FALSE)
message(paste("Saved data to CSV:", csv_full_path))

# PDF
file_name <- paste0("01.Feature Importance/Feature_Importance_Fold_", k, ".pdf")
full_path <- file.path(Outputs, file_name)
ggsave(
    full_path,  
    plot     = Plot,
    device = cairo_pdf,
    width    = 7, 
    height   = 7, 
    units    = "in")
message(paste("Saved graphic:", full_path))
}
```

#### Average of the folds

```{r}
library(shapviz)
library(dplyr)
library(ggplot2)
library(scales)

# Font
font <- "sans"
# font <- "Fira Sans"
# font <- "Source Sans Pro"
tamanio <- 17
all_importance_data <- list()

# ---------------------------------------------------------------------------------
# 1. LOOP TO ACCUMULATE IMPORTANCE SHAP OF THE V FOLDS
# ---------------------------------------------------------------------------------
for (k in 1:VFolds) {
  # 1. Load the final trained model for Fold k
  model_filename <- paste0("xgb-fit-final-model-fold-", k, ".rds")
  file_path <- file.path(Outputs, model_filename)
    
  # Handling the error if the model does not exist
  xgb.fit.final <- tryCatch({
      readRDS(file_path)
  }, error = function(e) {
      cat(paste0("Error: Could not load model for Fold ", k, ".\n"))
      return(NULL)
  })
  # Skip to the next fold if the model does not exist
  if (is.null(xgb.fit.final)) next 
  
  # 2. Get the training set from Fold k (features_train)
  current_data <- vtreat_processed_folds[[k]]
  features_train <- current_data$features_train
    
  # 3. Calculate the SHAP matrix for this Fold
  shp <- shapviz::shapviz(
        xgb.fit.final, 
        X_pred = features_train, 
        X = features_train)
  # 4. Extract the SHAP (Absolute Mean) Importance for this Fold
    importance_fold_k <- shapviz::sv_importance(
      shp, 
      kind = "bar", 
      max_display = 20L) %>%
        .$data %>%
        dplyr::mutate(Fold_ID = k)
    
  # Store in list
  all_importance_data[[k]] <- importance_fold_k
    
  print(paste("SHAP importance calculated for Fold", k))
}

# ---------------------------------------------------------------------------------
# # 2. CONSOLIDATION OF RESULTS AND FINAL GRAPH
# ---------------------------------------------------------------------------------
# 5. Consolidate all fold results
consolidated_importance <- dplyr::bind_rows(all_importance_data)

# 6. Calculate the Average SHAP Importance
importance_data_avg <- consolidated_importance %>%
    dplyr::group_by(feature) %>%
    dplyr::summarise(
        value = mean(value), # The average SHAP importance value
        .groups = 'drop'
    ) %>%
    dplyr::arrange(desc(value)) %>%
    dplyr::slice_head(n = 20) %>%
    # Join with the Labels dataframe to obtain the plot names
    dplyr::left_join(Labels, by = c("feature" = "Code_label")) %>%
    dplyr::select(Plot_label, value) %>%
    dplyr::rename(feature = Plot_label) %>%
    # Reorder the most important feature at the beginning
    dplyr::arrange(value)

# Factor the feature column (for ggplot order)
importance_data_avg$feature <- factor(importance_data_avg$feature, 
                                     levels = importance_data_avg$feature)

# ---------------------------------------------------------------------------------
# # 3. FINAL GRAPH
# ---------------------------------------------------------------------------------
Plot <- ggplot(importance_data_avg, aes(x = value, y = feature))+
        geom_segment(aes(x = 0, 
                         xend = value),
                     color = "#000000", 
                     linewidth = 0.5, 
                     linetype = "solid") +
        geom_point(aes(color = value > 0), 
                   size = 7, 
                   alpha = 1) +
        scale_color_manual(values = c("TRUE" = "#800080", "FALSE" = "#800080")) +
        labs(
          title = "A) Feature importance",
          x     = "Impact SHAP Value",
          y     = NULL)+
        theme_classic()+
        scale_x_continuous(
          limits = c(0, max(importance_data_avg$value) * 1.05), 
          expand = c(0, 0))+
        theme(
          plot.title = element_text(size   = tamanio, 
                                face   = "bold", 
                                hjust  = 0, 
                                color  = "#000000", 
                                family = font),
          plot.title.position = "plot",
          axis.title.x = element_text(size = tamanio),
          axis.text.x=element_text(size = tamanio,color = "#000000",family = font),
          axis.text.y=element_text(size = tamanio,color = "#000000",family = font),
          legend.position = "none",
          plot.margin = margin(5, 20, 5, 5)
    )
print(Plot)
# ---------------------------------------------------------------------------------
# 4. SAVING
# ---------------------------------------------------------------------------------
# CSV
importance_data_avg_for_csv <- importance_data_avg %>%
      dplyr::select(
        Feature_Label = feature,
        Impact_SHAP_Value = value)

importance_data_avg_for_csv <- importance_data_avg_for_csv %>%
    dplyr::arrange(dplyr::desc(Impact_SHAP_Value))
  
csv_file_name<- "01.Feature Importance/Feature_Importance_CV_Avg.csv"
csv_full_path <- file.path(Outputs, csv_file_name)
write.csv(importance_data_avg_for_csv, csv_full_path, row.names = FALSE)
message(paste("Saved data to CSV:", csv_full_path))

# PDF
full_path <- file.path(Outputs,"01.Feature Importance/Feature_Importance_CV_Avg.pdf")
ggsave(
    full_path,
    plot     = Plot, 
    device = cairo_pdf,
    width    = 7, 
    height   = 7, 
    units    = "in")
message(paste("Saved graphic:", full_path))
```

### 7.2.Shap plot

#### Fold by Fold

```{r}
library(shapviz)
library(dplyr)
library(ggplot2)
library(scales)
library(showtext)

# Font
font <- "sans"
# font <- "Fira Sans"
# font <- "Source Sans Pro"
tamanio <- 17

# ---------------------------------------------------------------------------------
# 0. PREPARATION
# ---------------------------------------------------------------------------------
for (k in 1:VFolds) {
  # 1. Load the final trained model for Fold k
  model_filename <- paste0("xgb-fit-final-model-fold-", k, ".rds")
  file_path <- file.path(Outputs, model_filename)
  
    # Handling the error if the model does not exist
  xgb.fit.final <- tryCatch({
      readRDS(file_path)
  }, error = function(e) {
      cat(paste0("Error: Could not load model for Fold ", k, ".\n"))
      return(NULL)
  })
  # Skip to the next fold if the model does not exist
  if (is.null(xgb.fit.final)) next 
  
  # 2. Get the training set from Fold k (features_train)
  current_data <- vtreat_processed_folds[[k]]
  features_train <- current_data$features_train
# ---------------------------------------------------------------------------------
# 1.UNIQUE SHAP CALCULATION
# ---------------------------------------------------------------------------------
shp <- shapviz(
    xgb.fit.final, 
    X_pred = features_train, 
    X = features_train      
)
## print(shp)
## print(names(shp$X))

# ---------------------------------------------------------------------------------
# 2.DATA EXTRACTION, LABELING AND ORDERING
# ---------------------------------------------------------------------------------
# Extract the 20 most important features, mapear las etiquetas
# y preparar el orden para el eje Y.
importance_data <- shapviz::sv_importance(shp, kind = "bee", max_display = 20L) %>%
    .$data %>%
    dplyr::distinct(feature) %>% # Get unique feature names 
    dplyr::rename(labels = feature) %>%
    # map the labels
    dplyr::left_join(Labels, by = c("labels" = "Code_label")) %>%
    dplyr::rename(feature = Plot_label) %>%
    dplyr::select(feature)
  
# Factor the features column to ensure order
importance_data$feature <- factor(importance_data$feature, 
                                 levels = rev(importance_data$feature))

## print(importance_data)

# ---------------------------------------------------------------------------------
# 3.GRAPH GENERATION
# ---------------------------------------------------------------------------------
Plot <- shapviz::sv_importance(
  shp,
  kind = "bee",
  max_display = 20L,
  viridis_args = list(option = "A"), # "inferno"
  bee_width = 0.2)+ # Adjust the spread of the points
  
  # Y-axis Label Adjustment
  scale_y_discrete(labels = rev(importance_data$feature)) +
    
  # X-Axis Adjustment
  scale_x_continuous(
        breaks = scales::pretty_breaks(n = 5))+
  # Title
  ggtitle("B) SHAP Values") +

  # Custom Styles
  theme(
      axis.title.x = element_text(size = tamanio, family = font, face = "plain"),
      axis.text.y = element_text(size = tamanio, color = "#000000", family = font),
      axis.text.x = element_text(size = (tamanio - 2), color = "#000000", family = font),
      legend.title = element_text(size = tamanio, family = font, face = "plain"),
      legend.text = element_text(size = tamanio, family = font, face = "plain"),
  # Grid lines
      panel.grid.major.x = element_line(color = "#e0e0e0", linetype = "dotted"),
  # Axes and Margins
      axis.line = element_line(color = "black", linewidth = 0.8),
      plot.title=element_text(size=tamanio,family=font, hjust = 0, face = "bold"),
      plot.title.position = "plot",
      panel.background = element_rect(fill = "transparent")
    )

print(Plot)
# ---------------------------------------------------------------------------------
# 4. SAVING
# ---------------------------------------------------------------------------------
file_name <- paste0("02.Shap Plot/Shap_Plot_Fold_", k, ".pdf")
full_path <- file.path(Outputs, file_name)
ggsave(
    full_path, 
    plot     = Plot, 
    device = cairo_pdf,
    width    = 7, 
    height   = 7, 
    units = "in", 
    bg = "white") 
message(paste("Saved graphic:", full_path))
}
```

#### Average of the folds

```{r}
library(shapviz)
library(dplyr)
library(ggplot2)
library(scales)

# Font
font <- "sans"
# font <- "Fira Sans"
# font <- "Source Sans Pro"
tamanio <- 17
all_shap_objects <- list()

# ---------------------------------------------------------------------------------
# 1. LOOP TO ACCUMULATE IMPORTANCE SHAP OF THE V FOLDS
# ---------------------------------------------------------------------------------
for (k in 1:VFolds) {
  # 1. Load the final trained model for Fold k
  model_filename <- paste0("xgb-fit-final-model-fold-", k, ".rds")
  file_path <- file.path(Outputs, model_filename)
    
  # Handling the error if the model does not exist
  xgb.fit.final <- tryCatch({
      readRDS(file_path)
  }, error = function(e) {
      cat(paste0("Error: Could not load model for Fold ", k, ".\n"))
      return(NULL)
  })
  # Skip to the next fold if the model does not exist
  if (is.null(xgb.fit.final)) next 
  
  # 2. Get the training set from Fold k (features_train)
  current_data <- vtreat_processed_folds[[k]]
  features_train <- current_data$features_train
    
  # 3. Calculate the SHAP matrix for this Fold
  shp <- shapviz::shapviz(
        xgb.fit.final, 
        X_pred = features_train, 
        X = features_train)
  
  # Store in list
  all_shap_objects[[k]] <- shp$feature_list
    
  message(paste("SHAP matrix calculated for Fold", k))
}
# ---------------------------------------------------------------------------------
# # 2. CONSOLIDATION OF RESULTS
# ---------------------------------------------------------------------------------
# 5. The SHAP matrices are added together and divided by V
avg_shap_matrix <- Reduce(`+`, all_shap_objects) / VFolds
shp_avg <- shp
shp_avg$feature_list <- avg_shap_matrix

# 6. Create the final SHAP object AVERAGE
importance_data <- shapviz::sv_importance(shp_avg, kind = "bee", max_display = 20L) %>%
    .$data %>%
    dplyr::distinct(feature) %>%
    dplyr::rename(labels = feature) %>%
    dplyr::left_join(Labels, by = c("labels" = "Code_label")) %>%
    dplyr::rename(feature = Plot_label) %>%
    dplyr::select(feature)

# Factor the feature column (for ggplot order)
importance_data_avg$feature <- factor(importance_data_avg$feature, 
                                     levels = importance_data_avg$feature)

# ---------------------------------------------------------------------------------
# # 3. FINAL GRAPH
# ---------------------------------------------------------------------------------
Plot <- shapviz::sv_importance(
  shp_avg,
  kind = "bee",
  max_display = 20L,
  viridis_args = list(option = "A"), # "inferno"
  bee_width = 0.2)+ # Adjust the spread of the points
  
  # Y-axis Label Adjustment
  scale_y_discrete(labels = rev(importance_data$feature)) +
    
  # X-Axis Adjustment
  scale_x_continuous(
        breaks = scales::pretty_breaks(n = 5))+
  # Title
  ggtitle("B) SHAP Values") +

  # Custom Styles
  theme(
      axis.title.x = element_text(size = tamanio, family = font, face = "plain"),
      axis.text.y = element_text(size = tamanio, color = "#000000", family = font),
      axis.text.x = element_text(size = (tamanio - 2), color = "#000000", family = font),
      legend.title = element_text(size = tamanio, family = font, face = "plain"),
      legend.text = element_text(size = tamanio, family = font, face = "plain"),
  # Grid lines
      panel.grid.major.x = element_line(color = "#e0e0e0", linetype = "dotted"),
  # Axes and Margins
      axis.line = element_line(color = "black", linewidth = 0.8),
      plot.title=element_text(size=tamanio,family=font, hjust = 0, face = "bold"),
      plot.title.position = "plot",
      panel.background = element_rect(fill = "transparent")
    )

print(Plot)
# ---------------------------------------------------------------------------------
# 4. SAVING
# ---------------------------------------------------------------------------------
full_path <- file.path(Outputs,"02.Shap Plot/Shap_Plot_CV_Avg.pdf")
ggsave(
    full_path,
    plot     = Plot, 
    device   = cairo_pdf,
    width    = 7, 
    height   = 7, 
    units    = "in",
    bg = "white")
message(paste("Saved graphic:", full_path))
```

### 7.3.Shap Dependence

#### Fold by Fold

```{r}
library(shapviz)
library(dplyr)
library(ggplot2)
library(scales)
library(stringr)

# Font
font <- "sans"
# font <- "Fira Sans"
# font <- "Source Sans Pro"
tamanio <- 17
top_n <- 7

# ---------------------------------------------------------------------------------
# 0. PREPARATION
# ---------------------------------------------------------------------------------
for (k in 1:VFolds){
  # 1. Load the final trained model for Fold k
  model_filename <- paste0("xgb-fit-final-model-fold-", k, ".rds")
  file_path <- file.path(Outputs, model_filename)
  
    # Handling the error if the model does not exist
  xgb.fit.final <- tryCatch({
      readRDS(file_path)
  }, error = function(e) {
      cat(paste0("Error: Could not load model for Fold ", k, ".\n"))
      return(NULL)
  })
  # Skip to the next fold if the model does not exist
  if (is.null(xgb.fit.final)) next 
  
  # 2. Get the training set from Fold k (features_train)
  current_data <- vtreat_processed_folds[[k]]
  features_train <- current_data$features_train
  current_split <- blocked_folds$splits[[k]]
  train_sample <- rsample::training(current_split)
# ---------------------------------------------------------------------------------
# 1.UNIQUE SHAP CALCULATION
# ---------------------------------------------------------------------------------
shp <- shapviz(
    xgb.fit.final, 
    X_pred = features_train, 
    X = features_train      
)
## print(shp)
## print(names(shp$X))

# ---------------------------------------------------------------------------------
# 2.IDENTIFICATION OF TOP N FEATURES
# ---------------------------------------------------------------------------------
# 3. Extract the n most important features (by absolute mean SHAP value)
top_n_features <- shapviz::sv_importance(shp, kind = "bee", max_display = top_n) %>%
    .$data %>%
    dplyr::distinct(feature) %>% 
    dplyr::pull(feature) # Extract the code vector
  
# ---------------------------------------------------------------------------------
# 2.CREATE LONG SHAPS
# ---------------------------------------------------------------------------------
# 4. Generate the SHAP Long dataset
shap_long <- shap.prep(xgb_model = xgb.fit.final, X_train = features_train)

  #1.1 Area Code
  area_codes_key <- train_sample %>%
  dplyr::mutate(Area_code = as.integer(Area_code)) %>% 
  sf::st_drop_geometry() %>% 
  dplyr::select(Area_code) %>%
  dplyr::mutate(ID = row_number()) 
  
  selected_data <- shap_long %>%
  dplyr::filter(variable %in% top_n_features) %>% 
  dplyr::left_join(area_codes_key, by = "ID") %>% 
  dplyr::left_join(Labels, by = c("variable" = "Code_label")) %>%
  dplyr::select(-ID)%>% 
  dplyr::mutate(Fold_ID = k)


# The list of variables for the loop:
selected_variables <- unique(selected_data$variable)
# ---------------------------------------------------------------------------------
# 3. SAVING CSV
# ---------------------------------------------------------------------------------
 shap_data_for_csv <- selected_data %>%
      dplyr::select(
        Fold_ID,
        Area_code, 
        Feature = variable, 
        Feature_Label = Plot_label,
        Feature_Value = rfvalue, 
        SHAP_Value = value)
  
csv_file_name <- paste0("03.Shap Dependence/SHAP_Dependence_Fold_", k, ".csv")
csv_full_path <- file.path(Outputs, csv_file_name)
write.csv(shap_data_for_csv, csv_full_path, row.names = FALSE)
message(paste("Saved SHAP Dependence Data for Fold", k, "to CSV:", csv_full_path))
# ---------------------------------------------------------------------------------
# 3.GRAPH GENERATION
# ---------------------------------------------------------------------------------
for (var_code in selected_variables) {
  # 5. Filter data for the current variable
  data_subset <- selected_data %>% dplyr::filter(variable == var_code)
    
  # 6. Obtain legible title (Labeling)
  plot_title_label <- Labels %>%
      dplyr::filter(Code_label == var_code) %>%
      dplyr::pull(Plot_label)
    
  # If the label is not found, use the code
  if (length(plot_title_label) == 0) {
      plot_title_label <- var_code}
    
  # 7. Generation of the Dependency Graph
  Plot <- ggplot(data_subset, aes(x = rfvalue, y = value))+
    
    # Points (SHAP value vs. Feature value)
    geom_point(aes(color = rfvalue), alpha = 0.5, size = 3)+
    
    # Trend Line
    geom_smooth(method = "gam", formula = y ~ s(x, bs = "cs"), 
                    se = TRUE, fill = "grey90", color = "black", linewidth = 1.2)+
    
    # Styles and Scales
    scale_color_viridis_c(option = "inferno") +
    scale_y_continuous(labels = scales::label_number(scale_cut = scales::cut_short_scale(), accuracy = 0.001)) +
        
    # Titles
    labs(
        #title = plot_title_label, 
        x = "Feature Value", 
        y = "SHAP Value") +
    theme_classic()+
        
    # Custom Styles
    theme(
#plot.title = element_text(size = 20, family = font, face = "bold", hjust = 0)
axis.title.x = element_text(family = font, size = 25, color = "black", margin = margin(t = 10)),
axis.title.y = element_text(family = font, size = 25, color = "black", margin = margin(r = 10)),
axis.text.x  = element_text(family = font, size = 17, color = "black"),
axis.text.y  = element_text(family = font, size = 17, color = "black"),

legend.position = "none",
axis.line = element_line(color = "black", linewidth = 0.8),
plot.margin = margin(10, 20, 10, 10)
)
  
print(Plot)
# ---------------------------------------------------------------------------------
# 4. SAVING
# ---------------------------------------------------------------------------------
# PDF
sanitized_title <- str_replace_all(plot_title_label, "[^a-zA-Z0-9_.]", "_")
file_name <- paste0("03.Shap Dependence/Shap_Dependence_Fold_",k,"_", sanitized_title, ".pdf")
full_path <- file.path(Outputs, file_name)  
ggsave(full_path, 
       plot   = Plot, 
       device = cairo_pdf,
       width  = 10, 
       height = 7, 
       units = "in", 
       bg = "white") 
message(paste("Saved graphic:", full_path))
}
}
```

#### Average of the folds

```{r}
library(shapviz)
library(dplyr)
library(ggplot2)
library(scales)
library(stringr)

# Font
font <- "sans"
# font <- "Fira Sans"
# font <- "Source Sans Pro"
tamanio <- 17
top_n <- 7

all_shap_long_data <- list()
# ---------------------------------------------------------------------------------
# 1. LOOP TO ACCUMULATE IMPORTANCE SHAP OF THE V FOLDS
# ---------------------------------------------------------------------------------
for (k in 1:VFolds) {
  # 1. Load the final trained model for Fold k
  model_filename <- paste0("xgb-fit-final-model-fold-", k, ".rds")
  file_path <- file.path(Outputs, model_filename)
    
  # Handling the error if the model does not exist
  xgb.fit.final <- tryCatch({
      readRDS(file_path)
  }, error = function(e) {
      cat(paste0("Error: Could not load model for Fold ", k, ".\n"))
      return(NULL)
  })
  # Skip to the next fold if the model does not exist
  if (is.null(xgb.fit.final)) next 
  
  # 2. Get the training set from Fold k (features_train)
  current_data <- vtreat_processed_folds[[k]]
  features_train <- current_data$features_train
  current_split <- blocked_folds$splits[[k]]
  train_sample <- rsample::training(current_split)
    
  # 3. Calculate the SHAP matrix for this Fold
  shp <- shapviz::shapviz(
        xgb.fit.final, 
        X_pred = features_train, 
        X = features_train)
  
  # 4. Generate the SHAP Long dataset
    shap_long <- shap.prep(xgb_model = xgb.fit.final, X_train = features_train)
    
  # 4.1 Area Codes
    area_codes_df <- train_sample %>%
      sf::st_drop_geometry() %>% 
      dplyr::mutate(ID = row_number()) %>% 
      dplyr::select(ID, Area_code)
    
    shap_long <- shap_long %>%
        dplyr::left_join(area_codes_df, by = "ID") %>%
        dplyr::select(-ID)
    
  # 5. Accumulate the long SHAP data
    shap_long$Fold_ID <- k # Add fold identifier for traceability
    all_shap_long_data[[k]] <- shap_long
  
  # 6. Accumulate the individual SHAP values (for average importance)
  # We also need the raw SHAP matrix to identify the overall order
    importance <- shapviz::sv_importance(shp, kind = "bar", max_display = length(colnames(features_train))) %>%
        .$data %>% 
        dplyr::mutate(Fold_ID = k)
    
  # 7. Save the importance for the calculation of global Top N
  if (k == 1) {
      all_importance <- importance
  } else {
      all_importance <- dplyr::bind_rows(all_importance, importance)
  }
}
# ---------------------------------------------------------------------------------
# # 2. CONSOLIDATION OF RESULTS
# ---------------------------------------------------------------------------------
# 8. Consolidate the entire SHAP Long dataset (all points from all folds)
consolidated_shap_long <- dplyr::bind_rows(all_shap_long_data)

# 9. Identify the ORDER OF GLOBAL IMPORTANCE (Consistent Top N)
top_n_features_global <- all_importance %>%
    dplyr::group_by(feature) %>%
    dplyr::summarise(avg_value = mean(value), .groups = 'drop') %>%
    dplyr::arrange(desc(avg_value)) %>%
    dplyr::slice_head(n = top_n) %>%
    dplyr::pull(feature)

# 10. Filter the Consolidated Long SHAP Dataset
selected_data_global <- consolidated_shap_long %>%
    dplyr::filter(variable %in% top_n_features_global)%>%
    dplyr::left_join(Labels, by = c("variable" = "Code_label"))

# La lista de variables para el loop:
selected_variables_global <- unique(selected_data_global$variable)

# ---------------------------------------------------------------------------------
# 3. SAVING CSV
# ---------------------------------------------------------------------------------
shap_data_for_csv <- selected_data_global%>%
  dplyr::select(
        Fold_ID, 
        Area_code, 
        Feature = variable,
        Feature_Label = Plot_label,
        Feature_Value = rfvalue, 
        SHAP_Value = value, 
        Standardized_Shape_Value = stdfvalue,
        Base_Shap = mean_value)
csv_file_name<- "03.Shap Dependence/Shap_Dependence_CV_Avg.csv"
csv_full_path <- file.path(Outputs, csv_file_name)
write.csv(shap_data_for_csv, csv_full_path, row.names = FALSE)
message(paste("Saved data to CSV:", csv_full_path))
# ---------------------------------------------------------------------------------
# # 4. FINAL GRAPH
# ---------------------------------------------------------------------------------
for (var_code in selected_variables_global){
    # 11. Filter data for the current variable (del dataset global)
    data_subset <- selected_data_global %>% dplyr::filter(variable == var_code)
    
    # 12. Obtain legible title (Labeling)
    plot_title_label <- Labels %>%
        dplyr::filter(Code_label == var_code) %>%
        dplyr::pull(Plot_label)
    
    # If the label is not found, use the code
    if (length(plot_title_label) == 0) {plot_title_label <- var_code}

    # 13. Generation of the Dependency Graph
    Plot <- ggplot(data_subset, aes(x = rfvalue, y = value))+
        
    # Points (SHAP value vs. Feature value)
    geom_point(aes(color = rfvalue), alpha = 0.5, size = 3)+ 
        
    # Trend Line (The trend line is the average of the 5 folds)
    geom_smooth(method = "gam", formula = y ~ s(x, bs = "cs"), 
                    se = TRUE, fill = "grey90", color = "black", linewidth = 1.2)+
        
    # Styles and Scales
    scale_color_viridis_c(option = "inferno") +
    scale_y_continuous(labels = scales::label_number(scale_cut = scales::cut_short_scale(), accuracy = 0.001)) +
        
    # Titles
    labs(
        #title = plot_title_label, 
        x = "Feature Value", 
        y = "SHAP Value") +
    theme_classic()+
        
    # Custom Styles
    theme(
#plot.title = element_text(size = 20, family = font, face = "bold", hjust = 0)
axis.title.x = element_text(family = font, size = 25, color = "black", margin = margin(t = 10)),
axis.title.y = element_text(family = font, size = 25, color = "black", margin = margin(r = 10)),
axis.text.x  = element_text(family = font, size = 17, color = "black"),
axis.text.y  = element_text(family = font, size = 17, color = "black"),

legend.position = "none",
axis.line = element_line(color = "black", linewidth = 0.8),
plot.margin = margin(10, 20, 10, 10)
)
  
print(Plot)
# ---------------------------------------------------------------------------------
# 5. SAVING
# ---------------------------------------------------------------------------------
# PDF
sanitized_title <- str_replace_all(plot_title_label, "[^a-zA-Z0-9_.]", "_")
file_name <- paste0("03.Shap Dependence/Shap_Dependence_CV_Avg_", sanitized_title, ".pdf")
full_path <- file.path(Outputs,file_name)
ggsave(
    full_path,
    plot     = Plot, 
    device   = cairo_pdf,
    width    = 10, 
    height   = 7, 
    units    = "in",
    bg = "white")
message(paste("Saved graphic:", full_path))
}
```

### 7.4.Heatmap

### 7.5.Map

#### Fold by Fold

```{r}
library(shapviz)
library(dplyr)
library(tidyr)
library(ggplot2)
library(sf)
library(stringr)
library(units)

# Font
font <- "sans"
tamanio <- 15
# font <- "Fira Sans"
# font <- "Source Sans Pro"
top_n <- 7
Tema <- "mako"

# ---------------------------------------------------------------------------------
# 0. SPATIAL PREPARATION
# ---------------------------------------------------------------------------------
# Municipalities
Boundaries <- Geographical_areas %>%
  st_transform(crs = 6372) %>%
  st_simplify(preserveTopology = TRUE, dTolerance = 1000) %>%
  st_make_valid() %>% 
  mutate(Area_code = as.integer(Area_code))
  
# States
state_boundaries <- Geographical_areas_states %>%
  st_transform(crs = 6372) %>%
  st_simplify(preserveTopology = TRUE, dTolerance = 1000) %>%
  st_make_valid() %>% 
  mutate(Area_code = as.integer(Area_code))

# ---------------------------------------------------------------------------------
# 1. PREPARATION
# ---------------------------------------------------------------------------------
for (k in 1:VFolds) {
  
  print(paste("--- Generating SHAP Maps for Fold:", k, "---"))
    
  # 1. Load the final trained model for Fold k
  model_filename <- paste0("xgb-fit-final-model-fold-", k, ".rds")
  file_path <- file.path(Outputs, model_filename)
  
  # Handling the error if the model does not exist
  xgb.fit.final <- tryCatch({
      readRDS(file_path)
  }, error = function(e) {
      cat(paste0("Error: Could not load model for Fold ", k, ".\n"))
      return(NULL)
  })
  # Skip to the next fold if the model does not exist
  if (is.null(xgb.fit.final)) next 
  
  # 2. Get the training set from Fold k (features_train)
  current_data <- vtreat_processed_folds[[k]]
  features_train <- current_data$features_train
  current_split <- blocked_folds$splits[[k]]
  train_sample <- rsample::training(current_split)
  
# ---------------------------------------------------------------------------------
## 2. RAW SHAP CALCULATION AND DATA EXTRACTION
# ---------------------------------------------------------------------------------
  # Calculate the raw SHAP values (Matrix) and the SHAP object
  shp <- shapviz::shapviz(
    xgb.fit.final, 
    X_pred = features_train, 
    X = features_train)
  
  shap_values_raw <- shap.values(
    xgb_model = xgb.fit.final, 
    X_train = features_train)
  
  aux_shap_scores <- as.data.frame(shap_values_raw$shap_score)
    
  # Extract Area_code and join (using the unprocessed train_sample)
  area_codes_df <- train_sample %>%
    sf::st_drop_geometry() %>% 
    dplyr::select(Area_code)
  aux_wide <- cbind(area_codes_df, aux_shap_scores) 
  
# ---------------------------------------------------------------------------------
# 3. GENERATION OF THE LONG DATAFRAME FOR MAPPING
# ---------------------------------------------------------------------------------   # Generate the input_df_with_shap_values ​​for the Fold k
  input_df_with_shap_values <- aux_wide %>%
    tidyr::pivot_longer(
        cols = -Area_code,
        names_to = "feature",
        values_to = "shap_value"
    ) %>%
    dplyr::mutate(Fold_ID = k) %>%
    dplyr::left_join(Labels, by = c("feature" = "Code_label")) %>%
    dplyr::select(Fold_ID, Area_code, feature, Plot_label,shap_value)
# ---------------------------------------------------------------------------------
# 4. IDENTIFICATION OF TOP N FEATURES
# ---------------------------------------------------------------------------------
# Extract the codes of the Top N most important features (by their absolute mean SHAP value)
top_n_features_codes <- shapviz::sv_importance(shp, kind = "bee", max_display = top_n) %>%
  .$data %>%
  dplyr::distinct(feature) %>%
  dplyr::pull(feature)

# Label Preparation and Mapping
top_n_labels_map <- Labels %>%
  dplyr::filter(Code_label %in% top_n_features_codes) %>%
  dplyr::select(Code_label, Plot_label)

# ---------------------------------------------------------------------------------
# 5. SAVING CSV
# ---------------------------------------------------------------------------------
  shap_data_for_csv <- input_df_with_shap_values %>%
      dplyr::filter(feature %in% top_n_features_codes)%>%
      dplyr::select(
        Fold_ID,
        Area_code, 
        Feature = feature, 
        Plot_label,
        SHAP_Value = shap_value)
  
  csv_file_name <- paste0("05.Map/Map_Fold_",k,".csv")
  csv_full_path <- file.path(Outputs, csv_file_name)
  write.csv(shap_data_for_csv,csv_full_path, row.names = FALSE)
  message(paste("Saved SHAP Map Data for Fold", k, "to CSV:", csv_full_path))
# ---------------------------------------------------------------------------------
# 6.GRAPH GENERATION
# ---------------------------------------------------------------------------------
for (feature_code in top_n_features_codes){
    
    # Get the map title
    plot_title_label <- top_n_labels_map %>%
      dplyr::filter(Code_label == feature_code) %>%
      dplyr::pull(Plot_label)
      
    # Prepare SHAP data: Filter the SHAP values dataframe
    shap_data_for_map <- input_df_with_shap_values %>%
      dplyr::filter(feature == feature_code) %>%
      dplyr::select(Area_code, shap_value)
    
    # Join the geometries with the SHAP values of the current feature
    map_data <- Boundaries %>%
      dplyr::left_join(shap_data_for_map, by = "Area_code")
      
    # Creation of the Choropleth Map
    Plot <- ggplot(data = map_data)+
    
    # Municipalities layer: Filled with SHAP values
    geom_sf(aes(fill = shap_value), color = NA, linewidth = 0.1) +
      
    # State layer: contours
    geom_sf(data = state_boundaries, fill = NA, color = "black", linewidth = 0.1) +
      
    # Stylized
    scale_fill_viridis_c(option = Tema, name = "SHAP Value", na.value = "grey80") +
      theme_void()+
      
    # Custom Styles
    theme(
        legend.position = c(0.0, 0.2),
        legend.justification = "left",
        legend.direction = "horizontal",
        legend.title = element_text(family = font, size = tamanio, face = "plain", hjust = 0.5),
        legend.text = element_text(family = font, size = 12),
        legend.key.width = unit(1.4, "cm"),
        legend.key.height = unit(0.5, "cm"),
        plot.margin = margin(10, 10, 10, 10)
      )
    
print(Plot)

# ---------------------------------------------------------------------------------
# 7. SAVING
# ---------------------------------------------------------------------------------
# PDF
sanitized_title <- str_replace_all(plot_title_label, "[^a-zA-Z0-9_.]", "_")
file_name <- paste0("05.Map/Map_Fold_",k,"_", sanitized_title, ".pdf")
full_path <- file.path(Outputs, file_name)  
ggsave(full_path, 
       plot   = Plot, 
       device = cairo_pdf,
       width  = 10, 
       height = 7, 
       units = "in", 
       bg = "white") 
message(paste("Saved graphic:", full_path))
}
} # END
```

#### Average of the folds

```{r}
library(shapviz)
library(dplyr)
library(tidyr)
library(ggplot2)
library(sf)
library(stringr)
library(units)

# Font
font <- "sans"
tamanio <- 15
# font <- "Fira Sans"
# font <- "Source Sans Pro"
top_n <- 7
Tema <- "mako"

# ---------------------------------------------------------------------------------
# 0. SPATIAL PREPARATION
# ---------------------------------------------------------------------------------
# Municipalities
Boundaries <- Geographical_areas %>%
  st_transform(crs = 6372) %>%
  st_simplify(preserveTopology = TRUE, dTolerance = 1000) %>%
  st_make_valid() %>% 
  mutate(Area_code = as.integer(Area_code))
  
# States
state_boundaries <- Geographical_areas_states %>%
  st_transform(crs = 6372) %>%
  st_simplify(preserveTopology = TRUE, dTolerance = 1000) %>%
  st_make_valid() %>% 
  mutate(Area_code = as.integer(Area_code))

all_shap_long_data <- list()
all_importance_cruda <- list()

# ---------------------------------------------------------------------------------
# 1. LOOP TO ACCUMULATE IMPORTANCE SHAP OF THE V FOLDS
# ---------------------------------------------------------------------------------
for (k in 1:VFolds) {
  # Load the final trained model for Fold k
  model_filename <- paste0("xgb-fit-final-model-fold-", k, ".rds")
  file_path <- file.path(Outputs, model_filename)
  
  # Handling the error if the model does not exist
  xgb.fit.final <- tryCatch({
      readRDS(file_path)
  }, error = function(e) {
      cat(paste0("Error: Could not load model for Fold ", k, ".\n"))
      return(NULL)
  })
  # Skip to the next fold if the model does not exist
  if (is.null(xgb.fit.final)) next 
  
  # Get the training set from Fold k (features_train)
  current_data <- vtreat_processed_folds[[k]]
  features_train <- current_data$features_train
  current_split <- blocked_folds$splits[[k]]
  train_sample <- rsample::training(current_split)
  
  # Calculate the raw SHAP values
  shap_values_raw <- shap.values(
    xgb_model = xgb.fit.final, 
    X_train = features_train)
  aux_shap_scores <- as.data.frame(shap_values_raw$shap_score)
    
  # Extract Area_code and join (using the unprocessed train_sample)
  area_codes_df <- train_sample %>%
    sf::st_drop_geometry() %>% 
    dplyr::select(Area_code)
  aux_wide <- cbind(area_codes_df, aux_shap_scores) 
  
  # Generate the input_df_with_shap_values ​​for the Fold k
  input_df_with_shap_values <- aux_wide %>%
    tidyr::pivot_longer(
        cols = -Area_code,
        names_to = "feature",
        values_to = "shap_value"
    ) %>%
    dplyr::mutate(Fold_ID = k)
    #dplyr::left_join(Labels, by = c("feature" = "Code_label")) %>%
    #dplyr::select(Area_code, feature, shap_value, Plot_label)
  
  all_shap_long_data[[k]] <- input_df_with_shap_values
  
  importance <- shapviz::sv_importance(shapviz::shapviz(
    xgb.fit.final, 
    X_pred = features_train, 
    X = features_train), 
    kind = "bar") %>%
        .$data %>% 
        dplyr::mutate(Fold_ID = k)
        
    all_importance_cruda[[k]] <- importance
    
    print(paste("SHAP data extracted for Fold", k))
}    
# ---------------------------------------------------------------------------------
# # 2. CONSOLIDATION OF RESULTS
# ---------------------------------------------------------------------------------
# Consolidate all SHAP Long (points) dataframes
consolidated_shap_long <- dplyr::bind_rows(all_shap_long_data)  
    
# Consolidate the order of importance and calculate the global TOP N
all_importance_df <- dplyr::bind_rows(all_importance_cruda)   

# Identify the ORDER OF GLOBAL IMPORTANCE (Consistent Top N)
top_n_features_global <- all_importance %>%
    dplyr::group_by(feature) %>%
    dplyr::summarise(avg_value = mean(value), .groups = 'drop') %>%
    dplyr::arrange(desc(avg_value)) %>%
    dplyr::slice_head(n = top_n) %>%
    dplyr::pull(feature)

# Calculate the average SHAP Value per Feature and per Area_code
# This is the input_df_with_shap_values_AVG for the final mapping
input_df_with_shap_values_avg <- consolidated_shap_long %>%
    dplyr::group_by(Area_code, feature) %>%
    dplyr::summarise(
        shap_value_avg = mean(shap_value), # Calculate the average SHAP value
        .groups = 'drop'
    ) %>%
    # Filter only the Top N global features
    dplyr::filter(feature %in% top_n_features_codes) %>%
    # Join with the final LABELS
    dplyr::left_join(Labels, by = c("feature" = "Code_label")) %>%
    dplyr::select(Area_code, feature, shap_value = shap_value_avg, Plot_label)

top_n_labels_map_final <- input_df_with_shap_values_avg %>%
    dplyr::distinct(feature, Plot_label)

# ---------------------------------------------------------------------------------
# 3. SAVING CSV
# ---------------------------------------------------------------------------------
shap_data_for_csv <- input_df_with_shap_values_avg %>%
    dplyr::select(
        Area_code, 
        Feature = feature, 
        Plot_label,
        SHAP_Value_Avg = shap_value)
csv_file_name <- paste0("05.Map/Map_CV_Avg.csv")
csv_full_path <- file.path(Outputs, csv_file_name)
write.csv(shap_data_for_csv, csv_full_path, row.names = FALSE)
message(paste("Saved consolidated (averaged) SHAP Map Data to CSV:",csv_full_path))
# ---------------------------------------------------------------------------------
# # 4. FINAL GRAPH
# ---------------------------------------------------------------------------------
for (feature_code in top_n_features_codes){
    
    # Get the map title
    plot_title_label <- top_n_labels_map_final %>%
      dplyr::filter(feature == feature_code) %>%
      dplyr::pull(Plot_label)
      
    # Prepare SHAP data: Filter the SHAP values dataframe
    shap_data_for_map <- input_df_with_shap_values_avg %>%
      dplyr::filter(feature == feature_code) %>%
      dplyr::select(Area_code, shap_value)
    
    # Join the geometries with the SHAP values of the current feature
    map_data <- Boundaries %>%
      dplyr::left_join(shap_data_for_map, by = "Area_code")
      
    # Creation of the Choropleth Map
    Plot <- ggplot(data = map_data)+
    
    # Municipalities layer: Filled with SHAP values
    geom_sf(aes(fill = shap_value), color = NA, linewidth = 0.1) +
      
    # State layer: contours
    geom_sf(data = state_boundaries, fill = NA, color = "black", linewidth = 0.1) +
      
    # Stylized
    scale_fill_viridis_c(option = Tema, name = "SHAP Value", na.value = "grey80") +
      theme_void()+
      
    # Custom Styles
    theme(
        legend.position = c(0.0, 0.2),
        legend.justification = "left",
        legend.direction = "horizontal",
        legend.title = element_text(family = font, size = tamanio, face = "plain", hjust = 0.5),
        legend.text = element_text(family = font, size = 12),
        legend.key.width = unit(1.4, "cm"),
        legend.key.height = unit(0.5, "cm"),
        plot.margin = margin(10, 10, 10, 10)
      )
    
print(Plot)
# ---------------------------------------------------------------------------------
# 5. SAVING PDF
# ---------------------------------------------------------------------------------
sanitized_title <- str_replace_all(plot_title_label, "[^a-zA-Z0-9_.]", "_")
file_name <- paste0("05.Map/Map_CV_Avg_", sanitized_title, ".pdf")
full_path <- file.path(Outputs,file_name)
ggsave(
    full_path,
    plot     = Plot, 
    device   = cairo_pdf,
    width    = 10, 
    height   = 7, 
    units    = "in",
    bg = "white")
message(paste("Saved graphic:", full_path))
}
```
