---
title: "census-variables"
format: html
editor: visual
---

```{r}
rm(list = ls())
```

```{r}
library(dplyr)

wd <- "/Volumes/DEBIAS"

# Output geography: "lsoa", "msoa" or "lad"
geography_level <- "lad"

if (!geography_level %in% c("lsoa", "msoa", "lad")) {
  stop("geography_level must be one of: 'lsoa', 'msoa', 'lad'")
}

geo_file_suffix <- ifelse(geography_level == "lad", "ltla", geography_level)

read_ts <- function(ts_id) {
  read.csv(
    paste0(
      wd,
      "/data/inputs/census/census2021-ts/",
      ts_id,
      "/",
      ts_id,
      "-",
      geo_file_suffix,
      ".csv"
    )
  )
}
```

# Read data and create census dataset

## Total resident population MSOA

```{r}

df_read <- read_ts("census2021-ts001") %>%
  select(date, geography, geography.code, Residence.type..Total..measures..Value) %>% rename("total_residents" = "Residence.type..Total..measures..Value")

df <- df_read
```

## Total households

```{r}
df_read <- read_ts("census2021-ts003")

df_read$total_hh <- df_read$Household.composition..Total..measures..Value

df_read <- df_read %>% select(geography.code, total_hh)

df <- df %>% left_join(df_read, 
                by = join_by(geography.code))
```

## Percentage of migrants

```{r}
df_read <- read_ts("census2021-ts004")

df_read$per_ukborn <- (df_read$Country.of.birth..Europe..United.Kingdom..measures..Value/df_read$Country.of.birth..Total..measures..Value) * 100

df_read$per_nonukborn <- 100 - df_read$per_ukborn

df_read <- df_read %>% select(geography.code, per_ukborn, per_nonukborn)

df <- df %>% left_join(df_read, 
                by = join_by(geography.code))
```

## Population density

```{r}
df_read <- read_ts("census2021-ts006")

df_read$pop_density <- df_read$Population.Density..Persons.per.square.kilometre..measures..Value

df_read <- df_read %>% select(geography.code, pop_density)

df <- df %>% left_join(df_read, 
                by = join_by(geography.code))
```

## Percentage of people in each 5-year age band

```{r}
df_read <- read_ts("census2021-ts007a")

for (i in 5:22) {
  column_name <- paste0("per_age_", (i-5)*5, "-", (i-5)*5+4)  # Generate column name
  df_read[[column_name]] <- df_read[[i]]/df_read[[4]]      # Assign values (for this example: multiples of i)
}

df_read <- df_read %>% select(geography.code, colnames(df_read)[23:40])

df_read[,2:19] <- df_read[,2:19] * 100

df_read <- df_read %>%
  mutate(
    `per_age_0-9`   = `per_age_0-4`  + `per_age_5-9`,
    `per_age_10-19` = `per_age_10-14` + `per_age_15-19`,
    `per_age_20-29` = `per_age_20-24` + `per_age_25-29`,
    `per_age_30-39` = `per_age_30-34` + `per_age_35-39`,
    `per_age_40-49` = `per_age_40-44` + `per_age_45-49`,
    `per_age_50-59` = `per_age_50-54` + `per_age_55-59`,
    `per_age_60-69` = `per_age_60-64` + `per_age_65-69`,
    `per_age_70plus` = `per_age_70-74` + `per_age_75-79` +
                     `per_age_80-84` + `per_age_85-89`
  ) %>%
  select(
    geography.code,
    `per_age_0-9`,
    `per_age_10-19`,
    `per_age_20-29`,
    `per_age_30-39`,
    `per_age_40-49`,
    `per_age_50-59`,
    `per_age_60-69`,
    `per_age_70plus`
  )

df <- df %>% left_join(df_read, 
                by = join_by(geography.code))
```

## Percentage of female

```{r}
df_read <- read_ts("census2021-ts008")

df_read$per_female <- (df_read$Sex..Female..measures..Value/df_read$Sex..All.persons..measures..Value) * 100

df_read <- df_read %>% select(geography.code, per_female)

df <- df %>% left_join(df_read, 
                by = join_by(geography.code))
```

## Household by deprivation dimensions (non-deprived households)

```{r}
df_read <- read_ts("census2021-ts011")

df_read$per_hh_notdeprived <- (df_read$Household.deprivation..Household.is.not.deprived.in.any.dimension..measures..Value/df_read$Household.deprivation..Total..All.households..measures..Value) * 100

df_read <- df_read %>% select(geography.code, per_hh_notdeprived)

df <- df %>% left_join(df_read, 
                by = join_by(geography.code))
```

## Year of arrival in UK

```{r}
# df_read <- read.csv(paste0(wd, "/data/inputs/census/census2021-ts/census2021-ts015/census2021-ts015-ltla.csv")) 
# 
# df_read$per_recent_migrant <- df_read$Year.of.arrival.in.the.UK..Arrived.2020.to.2021..measures..Value/df$total_residents
# 
# df_read <- df_read %>% select(geography.code, per_recent_migrant)
# 
# df <- df %>% left_join(df_read,
#                 by = join_by(geography.code))
```

## Length of residence

```{r}
df_read <- read_ts("census2021-ts016")

df_read$per_recent_migrant <- (df_read$Length.of.residence.in.the.UK..Less.than.2.years..measures..Value/df$total_residents ) * 100

df_read <- df_read %>% select(geography.code, per_recent_migrant)

df <- df %>% left_join(df_read,
                by = join_by(geography.code))

```

## Household size

```{r}
df_read <- read_ts("census2021-ts017")

df_read$house_six <- (df_read$Household.size..6.people.in.household..measures..Value/df_read$Household.size..Total..All.household.spaces..measures..Value ) * 100

df_read$house_seven <- (df_read$Household.size..7.people.in.household..measures..Value/df_read$Household.size..Total..All.household.spaces..measures..Value ) * 100

df_read$house_eightplus <- (df_read$Household.size..8.or.more.people.in.household..measures..Value/df_read$Household.size..Total..All.household.spaces..measures..Value ) * 100

df_read$per_large_households <- df_read$house_six + df_read$house_seven + df_read$house_eightplus

df_read <- df_read %>% select(geography.code, per_large_households)

df <- df %>% left_join(df_read,
                by = join_by(geography.code))
```

## Ethnic group, non-white (white being UK's racial majority)

```{r}
df_read <- read_ts("census2021-ts021")

df_read$per_non_white <- (1 - df_read$Ethnic.group..White/df_read$Ethnic.group..Total..All.usual.residents) * 100

df_read <- df_read %>% select(geography.code, per_non_white)

df <- df %>% left_join(df_read,
                by = join_by(geography.code))
```

## General health, percentage of bad or very bad health

```{r}
df_read <- read_ts("census2021-ts037")

df_read$per_bad_health <- ((df_read$General.health..Bad.health + df_read$General.health..Very.bad.health)/df_read$General.health..Total..All.usual.residents) * 100

df_read <- df_read %>% select(geography.code, per_bad_health)

df <- df %>% left_join(df_read,
                by = join_by(geography.code))
```

## Disability, percentage of disabled with day-to-day activities limited a lot

```{r}
df_read <- read_ts("census2021-ts038")

df_read$per_severe_disability <- (df_read$Disability..Disabled.under.the.Equality.Act..Day.to.day.activities.limited.a.lot/df_read$Disability..Total..All.usual.residents) * 100

df_read <- df_read %>% select(geography.code, per_severe_disability)

df <- df %>% left_join(df_read,
                by = join_by(geography.code))
```

## Car availability, percentage households with no car or van

```{r}
df_read <- read_ts("census2021-ts045")

df_read$per_hh_no_car <- (df_read$Number.of.cars.or.vans..No.cars.or.vans.in.household/df_read$Number.of.cars.or.vans..Total..All.households) * 100

df_read <- df_read %>% select(geography.code, per_hh_no_car)

df <- df %>% left_join(df_read,
                by = join_by(geography.code))
```

## Central heating, percentage households with no central heating

```{r}
df_read <- read_ts("census2021-ts046")

df_read$per_hh_no_centralheat <- (df_read$Type.of.central.heating.in.household..No.central.heating/df_read$Type.of.central.heating.in.household..Total..All.households) * 100

df_read <- df_read %>% select(geography.code, per_hh_no_centralheat)

df <- df %>% left_join(df_read,
                by = join_by(geography.code))
```

## Tenure, percentage of owned households

```{r}
df_read <- read_ts("census2021-ts054")

df_read$per_hh_owned <- (df_read$Tenure.of.household..Owned/df_read$Tenure.of.household..Total..All.households) * 100

df_read <- df_read %>% select(geography.code, per_hh_owned)

df <- df %>% left_join(df_read,
                by = join_by(geography.code))
```

## Method of travel to work

```{r}
df_read <- read_ts("census2021-ts061")

df_read$per_home_work <- (df_read$Method.of.travel.to.workplace..Work.mainly.at.or.from.home/df_read$Method.of.travel.to.workplace..Total..All.usual.residents.aged.16.years.and.over.in.employmen) * 100

df_read <- df_read %>% select(geography.code, per_home_work)

df <- df %>% left_join(df_read,
                by = join_by(geography.code))
```

## NS-SeC, percentage of adults over 16 by employment category (inc full-time students)

```{r}
df_read <- read_ts("census2021-ts062")


for (i in 5:13) {
  column_name <- paste0("per_", colnames(df_read)[i])
  df_read[[column_name]] <- df_read[[i]]/df_read[[4]]
}

df_read <- df_read %>% select(geography.code, colnames(df_read)[14:22])

df_read[, 2:10] <- df_read[, 2:10] * 100

df <- df %>% left_join(df_read,
                by = join_by(geography.code)) %>%
  rename("per_NS_SeC_L123_higher_managerial_administrative_professional" = "per_National.Statistics.Socio.economic.Classification..NS.SEC...L1..L2.and.L3.Higher.managerial..administrative.and.professional.occupations",
         "per_NS_SeC_L456_lower_managerial_administrative_professional" = "per_National.Statistics.Socio.economic.Classification..NS.SEC...L4..L5.and.L6.Lower.managerial..administrative.and.professional.occupations",
         "per_NS_SeC_L7_intermediate" = "per_National.Statistics.Socio.economic.Classification..NS.SEC...L7.Intermediate.occupations",
         "per_NS_SeC_L89_small_employers_own_account" = "per_National.Statistics.Socio.economic.Classification..NS.SEC...L8.and.L9.Small.employers.and.own.account.workers",
         "per_NS_SeC_L1011_lower_supervisory_technical" = "per_National.Statistics.Socio.economic.Classification..NS.SEC...L10.and.L11.Lower.supervisory.and.technical.occupations",
         "per_NS_SeC_L12_semi_routine" = "per_National.Statistics.Socio.economic.Classification..NS.SEC...L12.Semi.routine.occupations",
         "per_NS_SeC_L13_routine" = "per_National.Statistics.Socio.economic.Classification..NS.SEC...L13.Routine.occupations",
         "per_NS_SeC_L141142_never_worked_unemployed" = "per_National.Statistics.Socio.economic.Classification..NS.SEC...L14.1.and.L14.2.Never.worked.and.long.term.unemployed",
         "per_NS_SeC_L15_ft_students" = "per_National.Statistics.Socio.economic.Classification..NS.SEC...L15.Full.time.students",
         )


```

## Highest qual, percentage of adults over 16 by qualification (no qualification or \> level 4)

```{r}
df_read <- read_ts("census2021-ts067")

df_read$per_no_qualifications <- ( df_read$Highest.level.of.qualification..No.qualifications/df_read$Highest.level.of.qualification..Total..All.usual.residents.aged.16.years.and.over) * 100

df_read <- df_read %>% select(geography.code, per_no_qualifications)

df <- df %>% left_join(df_read,
                by = join_by(geography.code))


df_read <- read_ts("census2021-ts067")

df_read$per_level4 <- (df_read$
Highest.level.of.qualification..Level.4.qualifications.and.above/df_read$Highest.level.of.qualification..Total..All.usual.residents.aged.16.years.and.over) * 100

df_read <- df_read %>% select(geography.code, per_level4)

df <- df %>% left_join(df_read,
                by = join_by(geography.code))

```

## Rural population

```{r}
# 1. Load Census population by OA
pop_oa <- read.csv(paste0(wd, "/data/inputs/census/census2021-ts/census2021-ts001/census2021-ts001-oa.csv")) %>%
  select(geography.code, population = `Residence.type..Total..measures..Value`)

# 2. Load Rural–Urban Classification
ruc <- read.csv(paste0(wd, "/data/inputs/ons_rural-urban-classification/Rural_Urban_Classification_(2021)_of_Output_Areas_in_EW.csv")) %>%
  select(OA21CD, Urban_rural_flag)

# 3. Load OA to X lookup
oa_x <- read.csv(paste0(wd, "/data/inputs/ons_rural-urban-classification/Output_Area_to_Lower_layer_Super_Output_Area_to_Middle_layer_Super_Output_Area_to_Local_Authority_District_(December_2021)_Lookup_in_England_and_Wales_v3.csv")) %>%
  select(
    OA21CD,
    LSOA21CD,
    tidyselect::matches("^LSOA[0-9]{2}NM$"),
    MSOA21CD,
    tidyselect::matches("^MSOA[0-9]{2}NM$"),
    tidyselect::matches("^LAD[0-9]{2}CD$"),
    tidyselect::matches("^LAD[0-9]{2}NM$")
  )

# 4. Merge datasets
data <- pop_oa %>%
  left_join(ruc, join_by("geography.code" == "OA21CD")) %>%
  left_join(oa_x, join_by("geography.code" == "OA21CD"))

# 5. Define rural categories
rural_categories <- c(
  "Rural"
)

# 6. Compute rural percentage by selected geography
if (geography_level == "lsoa") {
  group_code_col <- "LSOA21CD"
  group_name_col <- grep("^LSOA[0-9]{2}NM$", names(data), value = TRUE)[1]
} else if (geography_level == "msoa") {
  group_code_col <- "MSOA21CD"
  group_name_col <- grep("^MSOA[0-9]{2}NM$", names(data), value = TRUE)[1]
} else {
  group_code_col <- grep("^LAD[0-9]{2}CD$", names(data), value = TRUE)[1]
  group_name_col <- grep("^LAD[0-9]{2}NM$", names(data), value = TRUE)[1]
}

if (is.na(group_name_col)) {
  data$geo_name_tmp <- data[[group_code_col]]
  group_name_col <- "geo_name_tmp"
}

rural_pct <- data %>%
  group_by(.data[[group_code_col]], .data[[group_name_col]]) %>%
  summarise(
    total_pop = sum(population, na.rm = TRUE),
    rural_pop = sum(population[Urban_rural_flag %in% rural_categories], na.rm = TRUE),
    rural_pct = 100 * rural_pop / total_pop
  ) %>%
  arrange(desc(rural_pct))

# View results
head(rural_pct)

rural_pct <- rural_pct %>%
  select(
    geography.code = all_of(group_code_col),
    rural_pop,
    rural_pct
  )

df <- df %>% left_join(rural_pct,
                by = "geography.code")
```

# Save data

```{r}
output_path <- paste0(
  wd,
  "/data/inputs/census/census2021-ts/combined-data-",
  geo_file_suffix,
  ".csv"
)
write.csv(df, output_path, row.names = FALSE)

if (geography_level == "lad") {
  write.csv(df, paste0(wd, "/data/inputs/census/census2021-ts/combined-data-lad.csv"), row.names = FALSE)
}
```
