---
title: "explain-bias"
format: html
editor: visual
---

# Installing / loading packages

First, clear workspace:

```{r}
rm(list=ls())
```

```{r}
wd <- "/Volumes/DEBIAS"
```

```{r}
#| warning: false
#| message: false
# library necessary packages
library(xgboost)
library(parallel)
library(recommenderlab)
library(doParallel)
library(rsample)
library(tidyverse)
library(sf)
library(moments)
library(SHAPforxgboost)
library(shapviz)
library(pdp)
library(rsample)
library(foreach)
library(patchwork)
library(viridis)
library(scales)
library(tidyplots)
library(spdep)
library(spatialsample)
```

# Set theme

```{r}
#| include = FALSE
# clear memory and adds tufte templates for plots and maps
source("../code/style/data-visualisation_theme.R")
```

# Data

## Geographies data import

```{r}
# Spatial model switches
include_spatial_lag_covariates <- FALSE
geo_level <- "lad"  # "lsoa", "msoa", "lad"
geo_file_suffix <- ifelse(geo_level == "lad", "ltla", geo_level)
geo_bias_suffix <- geo_level
lag_tag <- ifelse(include_spatial_lag_covariates, "lagcov", "nolagcov")
output_tag <- paste0("_", geo_level, "_spatialcv_", lag_tag)
VFolds <- 5
```

```{r}
get_boundary_path <- function(wd, geo_level) {
  boundary_root <- file.path(wd, "data/inputs/geographies/boundaries")
  if (geo_level == "lad") {
    file.path(boundary_root, "LAD_Dec_2021_GB_BFC_2022_simplified.gpkg")
  } else if (geo_level == "msoa") {
    file.path(boundary_root, "MSOA_Dec_2021_EW_BFC_2022_simplified.gpkg")
  } else if (geo_level == "lsoa") {
    file.path(boundary_root, "LSOA_Dec_2021_EW_BFC_2022_simplified.gpkg")
  } else {
    stop("Unknown geo_level value: ", geo_level)
  }
}

boundary_path <- get_boundary_path(wd, geo_level)

if (!file.exists(boundary_path)) {
  stop("Boundary file not found for geo_level='", geo_level, "': ", boundary_path)
}
```

```{r}
# administrative boundaries
lda_sdf <- st_read(boundary_path)
boundary_code_col <- if ("code" %in% names(lda_sdf)) {
  "code"
} else {
  grep("CD$", names(lda_sdf), value = TRUE)[1]
}
boundary_name_col <- if ("name" %in% names(lda_sdf)) {
  "name"
} else {
  grep("NM$", names(lda_sdf), value = TRUE)[1]
}
if (is.na(boundary_code_col)) {
  stop("Could not detect boundary code column in: ", boundary_path)
}
if (!is.na(boundary_name_col)) {
  lda_sdf <- lda_sdf %>% rename(code = all_of(boundary_code_col), name = all_of(boundary_name_col))
} else {
  lda_sdf <- lda_sdf %>% rename(code = all_of(boundary_code_col))
}
```

## Other data import

This first line defines the data source to be analysed. We have three options:

-   `twitter`

-   `fb_stt`

-   `fb_tts`

-   `mapp1`

-   `mapp2`

```{r}
dfd <- "mapp2"
```

```{r}
# Output helpers
get_model_output_dir <- function(dfd, wd) {
  if (dfd == "twitter") {
    file.path(wd, "data/outputs", dfd, "monthly/populations/explain-bias")
  } else if (dfd == "fb_tts") {
    file.path(wd, "data/outputs/fb/census-month/populations/tts/explain-bias")
  } else if (dfd == "fb_stt") {
    file.path(wd, "data/outputs/fb/census-month/populations/stt/explain-bias")
  } else if (dfd == "mapp1") {
    file.path(wd, "data/outputs/mapp1/census-month/populations/explain-bias")
  } else if (dfd == "mapp2") {
    file.path(wd, "data/outputs/mapp2/census-month/populations/explain-bias")
  } else {
    stop("Unknown dfd value: ", dfd)
  }
}

get_bias_input_dir <- function(dfd, wd) {
  if (dfd == "twitter") {
    file.path(wd, "data/outputs", dfd, "monthly/populations/active-population-bias")
  } else if (dfd == "fb_tts") {
    file.path(wd, "data/outputs/fb/census-month/populations/tts/active-population-bias")
  } else if (dfd == "fb_stt") {
    file.path(wd, "data/outputs/fb/census-month/populations/stt/active-population-bias")
  } else if (dfd == "mapp1") {
    file.path(wd, "data/outputs/mapp1/census-month/populations/active-population-bias")
  } else if (dfd == "mapp2") {
    file.path(wd, "data/outputs/mapp2/census-month/populations/active-population-bias")
  } else {
    stop("Unknown dfd value: ", dfd)
  }
}

get_census_covariate_path <- function(wd, geo_level) {
  base <- file.path(wd, "data/inputs/census/census2021-ts")
  if (geo_level == "lad") {
    p1 <- file.path(base, "combined-data-lad.csv")
    p2 <- file.path(base, "combined-data-ltla.csv")
    if (file.exists(p1)) {
      p1
    } else {
      p2
    }
  } else if (geo_level == "msoa") {
    file.path(base, "combined-data-msoa.csv")
  } else if (geo_level == "lsoa") {
    file.path(base, "combined-data-lsoa.csv")
  } else {
    stop("Unknown geo_level value: ", geo_level)
  }
}

model_output_dir <- get_model_output_dir(dfd, wd)
bias_input_dir <- get_bias_input_dir(dfd, wd)
covariate_path <- get_census_covariate_path(wd, geo_level)

if (!file.exists(covariate_path)) {
  stop("Covariate file not found for geo_level='", geo_level, "': ", covariate_path)
}

prelim_output_dir <- file.path("../outputs/preliminary", dfd, "explain-bias")
dir.create(prelim_output_dir, recursive = TRUE, showWarnings = FALSE)
```

The next code chunk can be run using spatial weights and nonspatial weights options:

-   distance band to contain at least one neighbour: `active_population_bias_<geo>_w_db.csv`

-   fixed optimal distance band based on GWR cross-validation: `active_population_bias_<geo>_w_fbw.csv`

-   k-nearest neighbour: `active_population_bias_<geo>_w_knn.csv`

-   queen: `active_population_bias_<geo>_w_queen.csv`

-   nonspatial: `active_population_bias_<geo>.csv`

```{r}
# Read in data

# census
df <- read.csv(covariate_path)

# census labels
df_var_lbl <- read_csv(paste0(wd,"/data/inputs/census/census2021-ts/dictionary.csv")) %>% 
  filter(!grepl("excluded", .$group, ignore.case = TRUE))

# If lag covariates are switched off, drop lag-related dictionary entries
if (!isTRUE(include_spatial_lag_covariates)) {
  df_var_lbl <- df_var_lbl %>%
    filter(!grepl("_lagq$", var_label5)) %>%
    filter(var_label5 != "bias_w")
}

```

```{r}
# digital footprint mobility data
df_bias <- read.csv(file.path(bias_input_dir, paste0("active_population_bias_", geo_bias_suffix, ".csv"))) %>%
  rename("geography.code" = "code")

if (geo_level == "lad" && "name" %in% names(df_bias) && dfd %in% c("twitter", "mapp1")) {
  df_bias <- dplyr::filter(df_bias, name != "City of London")
}

# add spatial weights to the label data frame if they are included 
if (isTRUE(include_spatial_lag_covariates) && "bias_w" %in% names(df_bias)) {
  df_var_lbl[46, ] <- list("bias_w", "spatial_lag_bias", "Spatial Lag", 
                      "spatial_lag_bias", "spatial_lag_bias", "bias_w")
}
```

## Data wrangling

### Remove areas in Scotland and Northern Ireland

```{r}
# from boundaries
lda_sdf <- lda_sdf %>%
  filter(!grepl("^[SN]", `code`))
```

## Select input predictors

```{r}
df <- df %>%
  select(all_of(df_var_lbl$var_label5))
```

### Extract centroids

```{r}
# extract centroids
centroids_df <- st_centroid(lda_sdf) %>% 
  st_coordinates() %>% 
  as.data.frame() %>% 
  rowid_to_column("row_id")

# transfer area ids
centroids_with_ids <- lda_sdf %>%
  st_set_geometry(NULL) %>%              # Remove geometry to avoid issues with join
  rowid_to_column("row_id") %>%         # Create a temporary row_id for merging
  cbind(centroids_df) 

# remove row_id and rename x,y
centroids_with_ids <- centroids_with_ids %>% 
  select(-c(row_id)) %>% 
  rename(
    longitude = "X",
    latitude = "Y"
  )

# join
df_extended <- full_join(df, centroids_with_ids, by = join_by(geography.code == code) ) 
```

If spatial weights are included as a feature:

```{r}
if (isTRUE(include_spatial_lag_covariates) && "bias_w" %in% names(df_bias)) {
  df_extended <- merge(df_extended, df_bias[, c("geography.code", "bias_w")], by = "geography.code")
}
```

```{r}
# Queen weights used for lagged covariates and residual Moran's I
nb_q <- spdep::poly2nb(lda_sdf, queen = TRUE)
w_matrix_q <- spdep::nb2listw(nb_q, style = "W", zero.policy = TRUE)

if (isTRUE(include_spatial_lag_covariates)) {
  # Ensure the same spatial order as LAD polygons when creating lagged covariates
  df_cov_ordered <- lda_sdf %>%
    st_set_geometry(NULL) %>%
    dplyr::transmute(geography.code = code) %>%
    dplyr::left_join(df, by = "geography.code")

  lag_candidates <- df_cov_ordered %>%
    dplyr::select(where(is.numeric)) %>%
    names() %>%
    setdiff(c("date")) %>%
    .[!grepl("date", ., ignore.case = TRUE)]

  lagged_covs <- purrr::map_dfc(lag_candidates, function(v) {
    tibble::tibble(
      !!paste0(v, "_lagq") := spdep::lag.listw(
        w_matrix_q, df_cov_ordered[[v]], NAOK = TRUE, zero.policy = TRUE
      )
    )
  })

  lagged_covs <- dplyr::bind_cols(df_cov_ordered["geography.code"], lagged_covs)
  df_extended <- dplyr::left_join(df_extended, lagged_covs, by = "geography.code")
}
```

```{r}
# Keep coordinates only for spatial block construction (not model inputs)
coords_df <- df_extended %>%
  dplyr::select(geography.code, longitude, latitude)
coords_crs <- sf::st_crs(lda_sdf)

# Standardize model covariates (exclude date and coordinates from inputs)
df_standardised <- df_extended %>%
  dplyr::select(-c(date, longitude, latitude)) %>% 
  mutate(across(where(is.numeric), ~ (. - mean(.)) / sd(.)))
```

```{r}
input_df <- merge(df_standardised, df_bias[, c("geography.code", "bias")], by = "geography.code")
```

```{r}
print(which(is.na(input_df), arr.ind = TRUE))

input_df <- input_df %>% 
  drop_na()
```

### Moran's I (Queen) for model covariates

```{r}
non_covariate_cols <- c("geography.code", "date", "geography", "bias", "longitude", "latitude")
covariate_cols <- setdiff(names(input_df), non_covariate_cols)

# Align covariates to the same polygon order used by nb_q/w_matrix_q
cov_aligned <- lda_sdf %>%
  sf::st_drop_geometry() %>%
  dplyr::transmute(geography.code = code) %>%
  dplyr::left_join(
    input_df %>% dplyr::select(geography.code, dplyr::all_of(covariate_cols)),
    by = "geography.code"
  )

covariate_moran_queen <- purrr::map_dfr(covariate_cols, function(v) {
  x_full <- cov_aligned[[v]]
  keep <- is.finite(x_full)

  nb_sub <- spdep::subset.nb(nb_q, subset = keep)
  lw_sub <- spdep::nb2listw(nb_sub, style = "W", zero.policy = TRUE)
  x_sub <- x_full[keep]

  mc <- spdep::moran.mc(
    x = x_sub,
    listw = lw_sub,
    nsim = 999,
    zero.policy = TRUE
  )

  tibble::tibble(
    variable = v,
    moran_i_queen = as.numeric(mc$statistic),
    p_value_queen = mc$p.value,
    n_used = length(x_sub)
  )
})

# Add readable labels where available
covariate_moran_queen <- covariate_moran_queen %>%
  dplyr::left_join(
    df_var_lbl %>% dplyr::select(var_label5, var_label2),
    by = c("variable" = "var_label5")
  ) %>%
  dplyr::mutate(variable_label = dplyr::coalesce(var_label2, variable)) %>%
  dplyr::select(variable, variable_label, moran_i_queen, p_value_queen, n_used) %>%
  dplyr::arrange(dplyr::desc(abs(moran_i_queen)))

readr::write_csv(
  covariate_moran_queen,
  file.path(prelim_output_dir, paste0("covariate-moran-queen", output_tag, ".csv"))
)

covariate_moran_queen
```

```{r}
rm(df_standardised, df_extended)
```

## Setting up model

```{r}
# Covariates
non_covariate_cols <- c("geography.code", "date", "geography", "bias", "longitude", "latitude")
xvars <- setdiff(names(input_df), non_covariate_cols)

# Final model inputs
y <- "bias"
x <- c(xvars)
```

```{r}
xvars
```

## Defining spatial blocked cross-validation folds

```{r}
set.seed(123)

# Add coordinates to input and convert to sf for blocked CV
input_sf <- input_df %>%
  dplyr::left_join(coords_df, by = "geography.code") %>%
  dplyr::mutate(.row_id = dplyr::row_number()) %>%
  sf::st_as_sf(coords = c("longitude", "latitude"), crs = coords_crs, remove = FALSE)

# Blocked CV partitions (spatialsample approach)
blocked_folds <- spatialsample::spatial_block_cv(
  input_sf,
  v = VFolds
)

# Use full data for model fitting; blocked folds are used for CV/tuning
train_sample <- input_sf %>% sf::st_drop_geometry()

# xgb.cv expects a list of assessment indices relative to training rows
spatial_folds <- lapply(blocked_folds$splits, function(s) rsample::assessment(s)$.row_id)
```

Features

```{r}
# variable names
features <- xvars

# Create the treatment plan from the training data
treatplan <- vtreat::designTreatmentsZ(train_sample, 
                                       features, 
                                       verbose = FALSE)

# Get the "clean" variable names from the scoreFrame
new_vars <- treatplan %>%
  magrittr::use_series(scoreFrame) %>%        
  dplyr::filter(code %in% c("clean", "lev")) %>% 
  magrittr::use_series(varName)  

# Prepare the training data
features_train <- vtreat::prepare(treatplan, 
                                  train_sample, 
                                  varRestriction = new_vars) %>% 
  as.matrix()

response_train <- train_sample[,y]

# dimensions of one-hot encoded data
dim(features_train)
```

# Training

## Exploring error trajectories

We run an initial XGBoost cross-validation using spatial blocked folds to inspect error trajectories. Hyperparameters are then tuned via grid search using the same blocked folds.

```{r}
# reproducibility
set.seed(123)

xgb.fit1 <- xgb.cv(
  data = features_train,
  label = response_train,
  nrounds = 1000,
  folds = spatial_folds,
  objective = "reg:squarederror",  # for regression models
  verbose = 0,               # silent,
#  early_stopping_rounds = 10
)
```

```{r}
# get number of trees that minimize error
xgb.fit1$evaluation_log %>%
  dplyr::summarise(
    ntrees.train = which(train_rmse_mean == min(train_rmse_mean))[1],
    rmse.train   = min(train_rmse_mean),
    ntrees.test  = which(test_rmse_mean == min(test_rmse_mean))[1],
    rmse.test   = min(test_rmse_mean),
  )

# plot error vs number trees
ggplot(xgb.fit1$evaluation_log) +
  geom_line(aes(iter, train_rmse_mean), color = "red") +
  geom_line(aes(iter, test_rmse_mean), color = "blue") +
  scale_x_log10() +
  scale_y_log10() 
```

Under spatial blocked cross-validation, both training and validation RMSE decrease sharply during the first boosting rounds and stabilise after a few trees. This indicates that most performance gains are achieved early, and adding many additional trees provides negligible improvement. We therefore use early stopping and select the number of trees at (or near) the minimum blocked-CV RMSE.

## Tuning

The tuned hyperparameters are:

-   eta: learning rate
-   max_depth: maximum tree depth
-   min_child_weight: minimum sum of instance weights in terminal nodes
-   subsample: fraction of observations sampled per tree
-   colsample_bytree: fraction of covariates sampled per tree
-   lambda: L2 regularisation
-   alpha: L1 regularisation

We then rank all configurations by blocked-CV RMSE and select the best-performing setting for the final model.

```{r}
# create hyperparameter grid (nrounds handled via early stopping, not tuned directly)
hyper_grid <- expand.grid(
  eta = c(0.03, 0.05, 0.1),            # Learning rate
  max_depth = c(2, 3, 5),              # Maximum depth of a tree
  min_child_weight = c(1, 3, 5),       # Minimum sum of instance weight in child
  subsample = c(0.7, 0.85, 1.0),       # Fraction of rows per tree
  colsample_bytree = c(0.7, 0.85, 1.0),# Fraction of covariates per tree
  lambda = c(0, 0.5, 1),               # L2 regularization (Ridge)
  alpha = c(0, 0.5, 1),                # L1 regularization (LASSO)
  optimal_trees = 0,                   # populated after CV
  min_RMSE = 0                         # populated after CV
)

nrow(hyper_grid)
```

## Parallelised code

We evaluate each hyperparameter combination using XGBoost under blocked spatial cross-validation and store, for each configuration, the minimum cross-validated RMSE and its corresponding optimal number of trees. This search is executed in parallel to reduce computation time

```{r}

# Measure the time taken for parallel grid search
time_taken <- system.time({
  
  # Set up parallel backend with the number of cores (leave one core free)
  num_cores <- max(1, detectCores() - 1)
  cl <- makeCluster(num_cores)
  registerDoParallel(cl)
  
  # Parallelized grid search
  results <- foreach(
    i = 1:nrow(hyper_grid),
    .combine = rbind,
    .packages = c("xgboost", "stats")
  ) %dopar% {
    
    # Create parameter list
    params <- list(
      eta = hyper_grid$eta[i],
      max_depth = hyper_grid$max_depth[i],
      min_child_weight = hyper_grid$min_child_weight[i],
      subsample = hyper_grid$subsample[i],
      colsample_bytree = hyper_grid$colsample_bytree[i],
      lambda = hyper_grid$lambda[i],
      alpha = hyper_grid$alpha[i]
    )
    
    # Reproducibility
    set.seed(123)
    
    # Train model with blocked spatial CV folds
    xgb.tune <- xgb.cv(
      params = params,
      data = features_train,
      label = response_train,
      nrounds = 500,
      folds = spatial_folds,
      objective = "reg:squarederror",
      verbose = 0,
      early_stopping_rounds = 20
    )
    
    # Collect optimal trees and minimum CV RMSE
    optimal_trees <- which.min(xgb.tune$evaluation_log$test_rmse_mean)
    min_RMSE <- min(xgb.tune$evaluation_log$test_rmse_mean)
    
    # Return one row per hyperparameter combination
    data.frame(
      eta = hyper_grid$eta[i],
      max_depth = hyper_grid$max_depth[i],
      min_child_weight = hyper_grid$min_child_weight[i],
      subsample = hyper_grid$subsample[i],
      colsample_bytree = hyper_grid$colsample_bytree[i],
      lambda = hyper_grid$lambda[i],
      alpha = hyper_grid$alpha[i],
      optimal_trees = optimal_trees,
      min_RMSE = min_RMSE
    )
  }
  
  # Convert to data frame if needed
  results <- as.data.frame(results)
  
  # Stop cluster
  stopCluster(cl)
  registerDoSEQ()
  
  # Display results
  print(results)
})

# Print elapsed time
print(time_taken)

```

## Non-parallelised code

::: callout-warning
This may take a while so leave it running overnight!
:::

```{r}

# # Measure the time taken for parallel grid search
# time_taken <- system.time({
#   
# # grid search 
# for(i in 1:nrow(hyper_grid)) {
# 
#   # create parameter list
#   params <- list(
#     eta = hyper_grid$eta[i],
#     max_depth = hyper_grid$max_depth[i],
#     min_child_weight = hyper_grid$min_child_weight[i],
#     subsample = hyper_grid$subsample[i],
#     colsample_bytree = hyper_grid$colsample_bytree[i]
#   )
# #
# #   # reproducibility
#   set.seed(123)
# 
#   # train model
#   xgb.tune <- xgb.cv(
#     params = params,
#     data = features_train,
#     label = response_train,
#     nrounds = 5000,
#     nfold = 10,
#     objective = "reg:squarederror",  # for regression models
#     verbose = 0,               # silent,
#     early_stopping_rounds = 10 # stop if no improvement for 10 consecutive trees
#   )
# #
# #   # add min training error and trees to grid
#    hyper_grid$optimal_trees[i] <- which.min(xgb.tune$evaluation_log$test_rmse_mean)
#    hyper_grid$min_RMSE[i] <- min(xgb.tune$evaluation_log$test_rmse_mean)
# }
# 
# })
# 
# # Print the time taken
# print(time_taken)
# 
# hyper_grid %>%
#   dplyr::arrange(min_RMSE) %>%
#   head(10)
# 
# print(hyper_grid)
```

```{r}
write_csv(results, file.path(model_output_dir, paste0("hypergrid", output_tag, ".csv")))
```

## Final model

Selecting optimal parameters

```{r}
hyper_grid <- read_csv(file.path(model_output_dir, paste0("hypergrid", output_tag, ".csv")))

optimal_pars <- hyper_grid %>% 
   slice_min(order_by = min_RMSE)
```

Once you've found the optimal model, we can fit our final model

```{r}
# parameter list
params <- list(
  eta = optimal_pars$eta[1],
  max_depth = optimal_pars$max_depth[1],
  min_child_weight = optimal_pars$min_child_weight[1],
  subsample = optimal_pars$subsample[1],
  colsample_bytree = optimal_pars$colsample_bytree[1],
  lambda = optimal_pars$lambda[1],
  alpha = optimal_pars$alpha[1]
)

# train final model using optimal number of trees from blocked CV
xgb.fit.final <- xgboost(
  params = params,
  data = features_train,
  label = response_train,
  nrounds = optimal_pars$optimal_trees[1],
  objective = "reg:squarederror",
  verbose = 0
)
```

Save optimal model

```{r}
saveRDS(xgb.fit.final, file.path(model_output_dir, paste0("xgb-fit-final", output_tag, ".rds")))
```

Extract parameters used for the final model

```{r}

write_csv(optimal_pars[1, ], file.path(prelim_output_dir, paste0("optimal_parameters", output_tag, ".csv")))
```

# Visualising

## Model performance

```{r}
# In-sample metrics (final model fitted on full data)
features_all <- vtreat::prepare(
  treatplan,
  train_sample,
  varRestriction = new_vars
) %>% as.matrix()

train_sample$bias_prediction <- predict(xgb.fit.final, features_all)

train_sample_rmse <- sqrt(mean((train_sample$bias - train_sample$bias_prediction)^2))
train_sample_pearson_cor <- cor(train_sample$bias, train_sample$bias_prediction, method = "pearson")
train_sample_spearman_cor <- cor(train_sample$bias, train_sample$bias_prediction, method = "spearman")
train_sample_sd_error <- sd(train_sample$bias - train_sample$bias_prediction)

# Blocked CV fold-level metrics using selected hyperparameters and optimal trees
cv_fold_metrics <- purrr::map_dfr(seq_along(blocked_folds$splits), function(k) {
  split_k <- blocked_folds$splits[[k]]
  fold_train <- rsample::analysis(split_k) %>% sf::st_drop_geometry()
  fold_test <- rsample::assessment(split_k) %>% sf::st_drop_geometry()

  treatplan_k <- vtreat::designTreatmentsZ(
    dframe = fold_train,
    varlist = xvars,
    verbose = FALSE
  )

  new_vars_k <- treatplan_k %>%
    magrittr::use_series(scoreFrame) %>%
    dplyr::filter(code %in% c("clean", "lev")) %>%
    magrittr::use_series(varName)

  x_train_k <- vtreat::prepare(treatplan_k, fold_train, varRestriction = new_vars_k) %>% as.matrix()
  y_train_k <- fold_train[[y]]
  x_test_k <- vtreat::prepare(treatplan_k, fold_test, varRestriction = new_vars_k) %>% as.matrix()
  y_test_k <- fold_test[[y]]

  model_k <- xgboost::xgboost(
    params = params,
    data = x_train_k,
    label = y_train_k,
    nrounds = optimal_pars$optimal_trees[1],
    objective = "reg:squarederror",
    verbose = 0
  )

  pred_k <- predict(model_k, x_test_k)
  resid_k <- y_test_k - pred_k

  tibble::tibble(
    fold = k,
    rmse = sqrt(mean(resid_k^2)),
    pearson = suppressWarnings(stats::cor(y_test_k, pred_k, method = "pearson")),
    spearman = suppressWarnings(stats::cor(y_test_k, pred_k, method = "spearman")),
    standard_dev = stats::sd(resid_k)
  )
})

write_csv(
  cv_fold_metrics,
  file.path(prelim_output_dir, paste0("model-performance-folds", output_tag, ".csv"))
)
```

```{r}
in_sample_metrics <- c(train_sample_rmse, train_sample_pearson_cor, train_sample_spearman_cor, train_sample_sd_error) %>% 
  round(3)

metric <- c("rmse", "pearson", "spearman", "standard_dev")

blocked_cv_metrics <- c(
  mean(cv_fold_metrics$rmse, na.rm = TRUE),
  mean(cv_fold_metrics$pearson, na.rm = TRUE),
  mean(cv_fold_metrics$spearman, na.rm = TRUE),
  mean(cv_fold_metrics$standard_dev, na.rm = TRUE)
) %>% round(3)

blocked_cv_sd <- c(
  sd(cv_fold_metrics$rmse, na.rm = TRUE),
  sd(cv_fold_metrics$pearson, na.rm = TRUE),
  sd(cv_fold_metrics$spearman, na.rm = TRUE),
  sd(cv_fold_metrics$standard_dev, na.rm = TRUE)
) %>% round(3)

summary_table <- data.frame(
  metric = metric,
  in_sample = in_sample_metrics,
  blocked_cv_mean = blocked_cv_metrics,
  blocked_cv_sd = blocked_cv_sd,
  optimal_trees = c(optimal_pars$optimal_trees[1], NA, NA, NA),
  data_source = dfd
)

summary_table
write_csv(summary_table, file.path(prelim_output_dir, paste0("model-performance", output_tag, ".csv")))
```

```{r}
# Predict + residuals
aux_full_sample <- vtreat::prepare(treatplan, input_df, varRestriction = new_vars) %>% as.matrix()
input_df$bias_prediction <- predict(xgb.fit.final, aux_full_sample)
input_df$model_residual <- input_df$bias - input_df$bias_prediction

# Collapse to one residual per area
resid_area <- input_df %>%
  dplyr::group_by(geography.code) %>%
  dplyr::summarise(model_residual = mean(model_residual, na.rm = TRUE), .groups = "drop")

# Align to boundary/weights order
resid_aligned <- lda_sdf %>%
  dplyr::filter(!grepl("^S", code)) %>%
  sf::st_set_geometry(NULL) %>%
  dplyr::transmute(geography.code = code) %>%
  dplyr::left_join(resid_area, by = "geography.code")

# Subset to valid residuals (must match nb_q length/order)
keep <- is.finite(resid_aligned$model_residual)
nb_sub <- spdep::subset.nb(nb_q, subset = keep)
lw_sub <- spdep::nb2listw(nb_sub, style = "W", zero.policy = TRUE)
x_sub <- resid_aligned$model_residual[keep]

# Moran's I
resid_moran <- spdep::moran.mc(
  x = x_sub,
  listw = lw_sub,
  nsim = 999,
  zero.policy = TRUE
)


model_residual_moran <- tibble::tibble(
  data_source = dfd,
  n_block_folds = VFolds,
  moran_i_residuals = as.numeric(resid_moran$statistic),
  p_value_residuals = resid_moran$p.value
)

write_csv(
  model_residual_moran,
  file.path(prelim_output_dir, paste0("model-residual-moran", output_tag, ".csv"))
)

model_residual_moran
```

## Variable importance

```{r}
# importance
importance_matrix <- xgb.importance(model = xgb.fit.final)

# dictionary with lag labels
df_dictionary_with_lags <- readr::read_csv(
  file.path(wd, "data/inputs/census/census2021-ts/dictionary_with_lags.csv"),
  show_col_types = FALSE
)

# map feature names to labels (fallback to raw name if no match)
importance_plot_df <- importance_matrix %>%
  dplyr::left_join(
    df_dictionary_with_lags %>% dplyr::select(var_label0, var_label2),
    by = c("Feature" = "var_label0")
  ) %>%
  dplyr::mutate(Feature_plot = dplyr::coalesce(var_label2, Feature)) %>%
  dplyr::select(Gain, Cover, Frequency, Feature = Feature_plot)

# plot with pretty labels
xgb.ggplot.importance(
  importance_plot_df,
  top_n = 30,
  measure = "Gain",
  xlab = "Relative importance"
) +
  scale_fill_viridis_d(name = "Cluster", option = "C", direction = 1) +
  theme_plot_tufte() +
  theme(
    plot.title = element_text(size = 20),
    legend.text = element_text(size = 14),
    legend.title = element_text(size = 16),
    axis.title.x = element_text(size = 18),
    axis.title.y = element_text(size = 18),
    panel.background = element_rect(
      fill = "grey85",
      colour = "grey85",
      linewidth = 0.5,
      linetype = "solid"
    )
  )
```

```{r}
png(file.path(prelim_output_dir, paste0("feature-importance_v2", output_tag, ".png")), units="in", width=8, height=10, res=300)
   last_plot()
dev.off()
```

## SHAP

Here is a gentle introduction to [SHAP values](https://blog.datascienceheroes.com/how-to-interpret-shap-values-in-r/).

Use of shapviz package <https://modeloriented.github.io/shapviz/>

Mean scores

```{r}
# Return the SHAP values and ranked features by mean|SHAP|
shap_values <- shap.values(xgb_model = xgb.fit.final, X_train = features_train)
# Ranked features by mean |SHAP|
# shap_values$mean_shap_score

shp <- shapviz(xgb.fit.final, X_pred = features_train, X = features_train)
```

**SHAP plots**

Bar plot - importance

```{r}
feature_lbl <- sv_importance(shp, kind = "bee", 
                             max_display = 20L) %>% 
  .$data %>% 
  distinct(.$feature) %>% 
  rename(labels = ".$feature") %>% 
  left_join(., df_dictionary_with_lags, join_by(labels == var_label0))

# Extract data from sv_importance plot
p_data <- sv_importance(shp, kind = "bar", 
                        viridis_args = list(option = "viridis"),
                        max_display = 20L)$data

min_val <- min(p_data$value, na.rm = TRUE)
max_val <- max(p_data$value, na.rm = TRUE)

sv_importance(shp, kind = "bar", 
              viridis_args = list(option = "viridis"),
              max_display = 20L) +
  scale_y_discrete( labels = rev(feature_lbl$var_label2)) +
  scale_x_continuous(
    labels = label_scientific(digits = 2),
    breaks = pretty_breaks(n = 3),
    expand = expansion(mult = c(0.02, 0.02))  # small padding left/right
  ) + 
  labs(
    x = "Mean SHAP value",
    y = "Features"
  ) +
  theme_plot_tufte() +
  theme(axis.title = element_text(size = 26),
        axis.text.y = element_text(size = 10),
        axis.text.x = element_text(size = 10))
```

```{r}
png(file.path(prelim_output_dir, paste0("feature-importance", output_tag, ".png")), units="in", width=8, height=10, res=300)
  last_plot()
dev.off()
```

### Extract data

```{r}
fi_data <- sv_importance(shp, kind = "bar", 
              viridis_args = list(option = "viridis"),
              max_display = 40L) %>% 
  .$data


write_csv(fi_data, file.path(prelim_output_dir, paste0("feature-importance-data", output_tag, ".csv")))
```

```{r}
feature_lbl <- sv_importance(shp, kind = "bee", 
                             max_display = 20L) %>% 
  .$data %>% 
  distinct(.$feature) %>% 
  rename(labels = ".$feature") %>% 
  left_join(., df_dictionary_with_lags, join_by(labels == var_label0))

sv_importance(shp, 
              kind = "bee", 
              max_display = 20L,
              viridis_args = list(option = "viridis") ) +
  scale_y_discrete( labels = rev(feature_lbl$var_label2)) +
  scale_x_continuous(
    labels = label_scientific(digits = 2),
    breaks = pretty_breaks(n = 3),
    expand = expansion(mult = c(0.02, 0.02))  # small padding left/right
  ) + 
  theme_plot_tufte() +
  theme(
    axis.title = element_text(size = 26),
    axis.text.y = element_text(size = 18),
    axis.text.x = element_text(size = 18),
    legend.title = element_text(size = 15),
    legend.text = element_text(size = 15))
  
```

### **How to interpret the plot above**

-   The y-axis indicates the variable name, in order of importance from top to bottom. The value next to them is the mean SHAP value.

-   The x-axis indicates the SHAP value and indicates how much is the change in log-odds.

-   The colour encodes the variable value.

-   Each point represents a row from the original dataset.

Lighter colours indicate that high values of a variable are associated with negative or positive values on the outcome / target variable.

Save shap plot

```{r}
png(file.path(prelim_output_dir, paste0("shap-plot-bias", output_tag, ".png")), units="in", width=8, height=10, res=300)
  last_plot()
dev.off()
```

## Partial dependence plots

Plots for selected variables

```{r}
# Prepare the long-format data:
shap_long <- shap.prep(xgb_model = xgb.fit.final, X_train = features_train)
# # is the same as: using given shap_contrib
shap_long <- shap.prep(shap_contrib = shap_values$shap_score, X_train = features_train)

# select variables
selected_data <- left_join(shap_long, head(feature_lbl, n = 8), 
                                join_by(variable == labels) ) %>% 
  dplyr::filter(!is.na(var_label2))

# select variables for plot
selected_variables <- unique(selected_data$var_label2)

# create a list to store individual plots
plot_list <- list()

# loop over each selected variable and create an individual plot
for (var in selected_variables) {
  # Filter data for the current variable
  data_subset <- selected_data %>% filter(var_label2 == var)
  
  # create a scatter plot for the current variable
  p <- ggplot(data_subset, aes(x = rfvalue, y = value, color = rfvalue)) +
    geom_smooth(color = "gray95", span = 0.3) +
    geom_point(alpha = 0.7, size = 3) +
    scale_colour_viridis_c(option = "D",
                           labels = label_number(scale_cut = cut_short_scale()),
                           guide = guide_colorbar(title.position="top")) +
  scale_y_continuous(
    labels = label_number(scale_cut = cut_short_scale(), accuracy = 0.001)
    ) +
    labs(title = var, x = "Feature value", y = "Shape value") +
    theme_plot_tufte() +
    theme(
    title = element_text(size = 16),
    legend.position = "none",
    text = element_text(size = 15)
    # legend.key.width = unit(2.5, "cm"), 
    # legend.key.height = unit(0.4, "cm"),
    # legend.text = element_text(size = 12)
  )
  
  # add the plot to the list
  plot_list[[var]] <- p
}

# combine all plots into a single layout
combined_plot <- wrap_plots(plot_list, ncol = 2)

# display combined plot
combined_plot
```

```{r}
png(file.path(prelim_output_dir, paste0("shap-dependence", output_tag, ".png")), units="in", width=8, height=10, res=300)
  last_plot()
dev.off()
```

::: callout-note
### **How to interpret the plot above**

The plot shows how a variable is associated with changes in shap values i.e. whether high score of a variable is associated with a decrease, increase or no change in shap values. They also provide a way to identify non-linear relationships.
:::

## Create a heatmap of LDAs based on feature importance

```{r}
xgb.fit.final_all <- xgboost(
  params = params,
  data = as.matrix(input_df[, xvars, drop = FALSE]),
  label = input_df$bias,
  nrounds = 2000,
  objective = "reg:squarederror",
  verbose = 0,
  early_stopping_rounds = 15)
```

```{r}
# Return the SHAP values and ranked features by mean|SHAP|
shap_values_full <- shap.values(xgb_model = xgb.fit.final_all, 
                                X_train = as.matrix(input_df[, xvars, drop = FALSE]))

aux_full <- as.data.frame(shap_values_full$shap_score)

input_df_with_shap_values <- cbind(input_df[,1:2], aux_full) %>% 
  pivot_longer(
    cols = c(3:length(.)),
    names_to = "feature",
    values_to = "shap_value"
  ) %>% 
  left_join(., df_dictionary_with_lags, join_by(feature == var_label5 ))
```

```{r}
selected_areas <- c(
  "Birmingham",
  "Leeds",
  "Sheffield",
  "Bradford",
  "Liverpool",
  "Manchester",
  "Bristol, City of",
  "Kirklees",
  "Croydon",
  "Barnet",
  "Isles of Scilly",
  "City of London",
  "Teesdale",
  "Berwick-upon-Tweed",
  "Alnwick",
  "Somerset West and Taunton",
  "Oswestry",
  "Rutland",
  "Shropshire",
  "Christchurch",
  "Purbeck",
  "Melton",
  "West Devon",
  "Ryedale",
  "Richmondshire",
  "Eden",
  "Castle Morpeth",
  "Corby"
)

# Ensure that var_label2 follows the same order as the bar chart
heatmap_labels <- feature_lbl$var_label2

input_df_with_shap_values %>% 
  filter(geography %in% selected_areas) %>% 
  filter(var_label2 %in% heatmap_labels) %>% 
  mutate(var_label2 = factor(var_label2, levels = rev(heatmap_labels))) %>%  # Ensure same order
  tidyplot(x = geography, 
           y = var_label2,  # Apply the consistent label order
           color = shap_value) %>%
  add_heatmap(rotate_labels = 90) %>% 
  adjust_size(height = 100) %>% 
  adjust_x_axis_title("") %>% 
  adjust_y_axis_title("") %>% 
  add(ggplot2::theme(
    legend.position = "bottom",
    legend.title = element_text(size=10, face = "plain", hjust=0.5, lineheight=0.45,
          color="black"),
    legend.key.width = unit(0.9, "cm"), 
    legend.key.height = unit(0.35, "cm"),
    legend.text = element_text(size = 8)
  )) %>% 
  add(ggplot2::labs(fill = "Shap value"))
```

```{r}

png(file.path(prelim_output_dir, paste0("heatmap_xxldas", output_tag, ".png")), units="in", width=8, height=10, res=300, bg = "transparent")
  last_plot()
dev.off()
```
